---
title: "Model workflow: inspect, edit, fit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model workflow: inspect, edit, fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ferx)
```

## Overview

This vignette covers two complementary workflows:

1. **Path-based workflow** — pass file paths directly to `ferx_fit()`. Straightforward for one-off fits.
2. **Pipe-based workflow** — bundle paths into a `ferx_model` object and chain steps with `|>`. Cleaner when you need to tweak options before fitting or chain post-fit steps.

Both workflows produce the same `ferx_fit` result.

**Entry point:** `ferx_model(model, data)` takes the model file first, following the convention of R modelling APIs like `lm(formula, data)`. The natural pipe entry is therefore the model:

```r
ferx_model(ex$model, data = ex$data) |> ferx_fit() |> summary()
```

If you prefer to start with the data — common in EDA pipelines — use the R 4.2+ `_` placeholder to route it to the named `data` argument:

```r
# Requires R >= 4.2
ex$data |> ferx_model(ex$model, data = _) |> ferx_fit() |> summary()
```

See [issue #81](https://github.com/InsightRX/ferx/issues/81) for the ongoing discussion on argument order.

---

## Path-based workflow

### Step 1 — Get a model file

`ferx_example()` returns paths to the bundled example models and their data:

```{r example}
ex <- ferx_example("warfarin")
ex$model  # path to warfarin.ferx
ex$data   # path to warfarin.csv
```

In a real workflow you would supply your own `.ferx` file here.

### Step 2 — Inspect before fitting

`ferx_model_inspect()` parses the model file and prints a concise summary of
its structure without running the optimizer. Use this to verify that ferx
interprets your model the way you expect before committing to a long run.

```{r inspect-pre}
ferx_model_inspect(ex$model)
```

The printed summary has four lines:

| Line | Meaning |
|------|---------|
| **Structural** | Detected PK model type and population parameter (theta) names |
| **IIV** | Inter-individual variability (omega) parameter names |
| **IOV** | Inter-occasion variability (kappa) parameter names, or `none` |
| **Residual** | Error model type (`proportional`, `additive`, or `combined`) |

`ferx_model_inspect()` returns the structure as a named list (invisibly), which
is useful for programmatic checks:

```{r inspect-programmatic}
s <- ferx_model_inspect(ex$model)
s$theta_names   # population parameter names
s$model_type    # short label for the PK model
s$iiv           # IIV parameter names
s$residual      # error model type
```

If the printed summary does not match your expectations — for example the model
type shows `NULL` or a parameter name is missing — fix the `.ferx` file before
proceeding. This catches typos and structural mistakes cheaply, before the
optimizer runs.

### Step 3 — Fit

Once the structure looks correct, pass the same paths to `ferx_fit()`:

```{r fit, eval = FALSE}
fit <- ferx_fit(ex$model, ex$data)
fit
```

### Step 4 — Re-inspect from the fitted result

After fitting you can call `ferx_model_inspect()` directly on the `ferx_fit`
object returned by `ferx_fit()` — no need to supply the model path again:

```{r inspect-post, eval = FALSE}
ferx_model_inspect(fit)
```

The output is identical to the pre-fit inspection. This is convenient when you
want to confirm the structure inside a script or report without keeping the
model path in scope.

---

## Pipe-based workflow

`ferx_model()` bundles model and data paths into a single object that flows
through a `|>` chain. This is the preferred style when you need to adjust
options before fitting or chain several post-fit steps.

### Entry point

```{r ferx-model}
ex <- ferx_example("warfarin")
m <- ferx_model(ex$model, data = ex$data)
m   # prints path, data path, and structure summary
```

### Minimal pipe — inspect, fit, summarise

`ferx_fit()` dispatches on the `ferx_model` object and picks up `$data`
automatically.

```{r minimal-pipe, eval = FALSE}
ferx_model(ex$model, data = ex$data) |>
  ferx_fit(method = "focei", covariance = TRUE) |>
  summary()
```

### Peek at a section mid-pipe

`ferx_get_section()` prints the named section to the console and passes the
`ferx_model` object through unchanged, so the pipe continues:

```{r get-section}
ferx_model(ex$model, data = ex$data) |>
  ferx_get_section("parameters")
```

```{r get-section-fit, eval = FALSE}
# Print [parameters] then fit without interrupting the chain
fit <- ferx_model(ex$model, data = ex$data) |>
  ferx_get_section("parameters") |>
  ferx_fit(method = "focei")
```

### Override fit options before fitting

`ferx_set_section()` rewrites a section on disk and returns the `ferx_model`
unchanged.
**Important:** it edits the file in place — copy first if you do not want to
modify the original.

```{r set-section, eval = FALSE}
model_copy <- file.path(tempdir(), "warfarin.ferx")
file.copy(ex$model, model_copy)

fit <- ferx_model(model_copy, data = ex$data) |>
  ferx_set_section("fit_options", c(
    "  method     = focei",
    "  maxiter    = 500",
    "  covariance = true"
  )) |>
  ferx_fit()

summary(fit)
ferx_model_inspect(fit)    # no path needed after fitting
ferx_cor_matrix(fit)       # parameter correlation matrix
ferx_plot_trace(fit)       # OFV + gradient norm convergence plot
```

### Validate before a long run

`ferx_check_init()` runs only 5 iterations and returns a trace and a summary
table. Use it to confirm the OFV is dropping before committing to a long
estimation:

```{r check-init, eval = FALSE}
chk <- ferx_check_init(ex$model, ex$data, method = "focei")
chk$summary      # ofv_start, ofv_end, ofv_drop — is OFV decreasing?
ferx_plot_trace(chk$fit)   # visual check: first few iterations
```

### Multi-stage method chain

Pass a character vector to `method` to chain stages — each stage starts from
the previous stage's converged parameters:

```{r method-chain, eval = FALSE}
fit <- ferx_model(ex$model, data = ex$data) |>
  ferx_fit(method = c("saem", "focei"), covariance = TRUE)
```

### Simulate and predict from fitted parameters

```{r sim-pred, eval = FALSE}
# Simulate 100 replicates using individual parameters from the fit
sim <- ferx_simulate(ex$model, ex$data, n_sim = 100, seed = 42, fit = fit)

# Population predictions using fitted theta (eta = 0)
pred <- ferx_predict(ex$model, ex$data, fit = fit)
```

### Post-fit diagnostics

All diagnostic outputs are fields on the `ferx_fit` object:

```{r diagnostics, eval = FALSE}
head(fit$sdtab)               # PRED, IPRED, CWRES, IWRES per observation
head(fit$ebe_etas)            # empirical Bayes ETAs per subject
head(fit$individual_estimates) # individual PK parameters (CL, V, KA, ...)
fit$eta_normality             # Shapiro-Wilk normality test per ETA
```

---

## Summary

### Path-based

```r
ex <- ferx_example("warfarin")
ferx_model_inspect(ex$model)          # verify structure before fitting
fit <- ferx_fit(ex$model, ex$data)
ferx_model_inspect(fit)               # re-inspect post-fit (no path needed)
```

### Pipe-based

```r
ex <- ferx_example("warfarin")

# Inspect, fit, summarise in one chain
ferx_model(ex$model, data = ex$data) |>
  ferx_get_section("parameters") |>  # optional: peek before fitting
  ferx_fit(method = "focei", covariance = TRUE) |>
  summary()

# Edit options then fit (copy model file first)
model_copy <- file.path(tempdir(), "warfarin.ferx")
file.copy(ex$model, model_copy)

ferx_model(model_copy, data = ex$data) |>
  ferx_set_section("fit_options", c(
    "  method     = focei",
    "  maxiter    = 500",
    "  covariance = true"
  )) |>
  ferx_fit() |>
  summary()
```
