---
title: "Editing ferx model files programmatically"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Editing ferx model files programmatically}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ferx)
```

ferx model files (`.ferx`) are plain text, so you can create and modify them
entirely from R without opening an editor. This vignette covers the three
functions that support that workflow: `ferx_model_new()`,
`ferx_model_section()`, and `ferx_model_set_section()`.

## Starting from scratch

`ferx_model_new()` writes a skeleton `.ferx` file for one of five built-in
templates: `"1cpt_oral"` (default), `"1cpt_iv"`, `"2cpt_oral"`, `"2cpt_iv"`,
and `"ode"`.

Pass `print = TRUE` to preview a template in the console without writing any
file — useful when you are exploring or copy-pasting a starting point:

```{r new-print}
ferx_model_new(print = TRUE)
```

```{r new-print-2cpt, eval = FALSE}
ferx_model_new(template = "2cpt_iv", print = TRUE)
```

To write a file, supply a path. By default the file is also opened in your
editor (`edit = TRUE`); pass `edit = FALSE` to skip that:

```{r new-write, eval = FALSE}
ferx_model_new("my_model.ferx", edit = FALSE)
```

## Inspecting a section

`ferx_model_section()` extracts the body of a named section and prints it.
Start with the bundled warfarin example:

```{r section}
ex <- ferx_example("warfarin")
ferx_model_section(ex$model, "parameters")
```

The return value is the character vector of lines, returned invisibly, so you
can capture it:

```{r section-capture}
params <- ferx_model_section(ex$model, "parameters")
params
```

## Scripted read-modify-write

The extract → modify → write-back pattern lets you change a model
programmatically without an editor. Here we raise the initial CL estimate in
the `parameters` section and write the result to a temporary copy:

```{r read-modify-write}
# work in a fresh temp file so re-knitting the vignette never reuses stale state
model_path <- tempfile(fileext = ".ferx")
stopifnot(file.copy(ex$model, model_path))

# read the section
lines <- ferx_model_section(model_path, "parameters")

# modify: change the TVCL initial value from 0.134 to 0.5
lines <- sub("TVCL\\([^,]+,", "TVCL(0.5,", lines)

# write back
ferx_model_set_section(model_path, "parameters", lines)

# verify
ferx_model_section(model_path, "parameters")
```

`ferx_model_set_section()` replaces only the targeted section; all other
sections are left untouched.

## Console-only workflow: create, tweak, fit

You can go from nothing to an estimated model without ever opening a text
editor. The steps below create a one-compartment oral model file, switch the
estimation method to FOCEI, and run `ferx_fit()`:

```{r console-only, eval = FALSE}
model_path <- file.path(tempdir(), "run1.ferx")

# 1. Write a skeleton
ferx_model_new(model_path, template = "1cpt_oral", edit = FALSE)

# 2. Switch method to focei
ferx_model_set_section(model_path, "fit_options", c(
  "  method     = focei",
  "  maxiter    = 300",
  "  covariance = true"
))

# 3. Fit (uses the bundled warfarin data for illustration)
ex   <- ferx_example("warfarin")
fit  <- ferx_fit(model_path, ex$data)
print(fit)
```

## Overwrite guards

By default `ferx_model_new()` refuses to overwrite an existing file:

```{r overwrite-error}
model_path <- tempfile(fileext = ".ferx")
ferx_model_new(model_path, edit = FALSE)            # creates the file

tryCatch(
  ferx_model_new(model_path, edit = FALSE),          # would overwrite — error
  error = function(e) message(conditionMessage(e))
)
```

Opt in explicitly with `overwrite = TRUE`:

```{r overwrite-ok}
ferx_model_new(model_path, template = "1cpt_iv", overwrite = TRUE, edit = FALSE)
ferx_model_section(model_path, "structural_model")
```
