---
title: "Model workflow: inspect, edit, fit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model workflow: inspect, edit, fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ferx)
```

## Overview

This vignette covers two complementary workflows:

1. **Path-based workflow** — pass file paths directly to `ferx_fit()`. Straightforward for one-off fits.
2. **Pipe-based workflow** — bundle paths into a `ferx_model` object and chain steps with `|>`. Cleaner when you need to tweak options before fitting or chain post-fit steps.

Both workflows produce the same `ferx_fit` result.

**Entry point:** `ferx_model(data, model)` takes the data file first, which lets a data object flow naturally into a pipeline:

```r
ex$data |> ferx_model(ex$model) |> ferx_fit() |> summary()
```

You can also call it positionally or by name:

```r
ferx_model(ex$data, ex$model) |> ferx_fit() |> summary()
ferx_model(data = ex$data, model = ex$model) |> ferx_fit() |> summary()
```

**Migration note:** Earlier versions used the order `ferx_model(model, data)`. Old-style positional calls (`ferx_model("pk.ferx")` or `ferx_model("pk.ferx", "data.csv")`) are detected by the `.ferx` extension and auto-corrected with a deprecation warning; this shim will be removed in a future release. Calls that pass `data` by name (`ferx_model("pk.ferx", data = "data.csv")`) keep working unchanged.

---

## Path-based workflow

### Step 1 — Get a model file

`ferx_example()` returns paths to the bundled example models and their data:

```{r example}
ex <- ferx_example("warfarin")
ex$model  # path to warfarin.ferx
ex$data   # path to warfarin.csv
```

In a real workflow you would supply your own `.ferx` file here.

### Step 2 — Inspect before fitting

`ferx_model_inspect()` parses the model file and prints a concise summary of
its structure without running the optimizer. Use this to verify that ferx
interprets your model the way you expect before committing to a long run.

```{r inspect-pre}
ferx_model_inspect(ex$model)
```

The printed summary has four lines:

| Line | Meaning |
|------|---------|
| **Structural** | Detected PK model type and population parameter (theta) names |
| **IIV** | Inter-individual variability (omega) parameter names |
| **IOV** | Inter-occasion variability (kappa) parameter names, or `none` |
| **Residual** | Error model type (`proportional`, `additive`, or `combined`) |

`ferx_model_inspect()` returns the structure as a named list (invisibly), which
is useful for programmatic checks:

```{r inspect-programmatic}
s <- ferx_model_inspect(ex$model)
s$theta_names   # population parameter names
s$model_type    # short label for the PK model
s$iiv           # IIV parameter names
s$residual      # error model type
```

If the printed summary does not match your expectations — for example the model
type shows `NULL` or a parameter name is missing — fix the `.ferx` file before
proceeding. This catches typos and structural mistakes cheaply, before the
optimizer runs.

### Step 3 — Fit

Once the structure looks correct, pass the same paths to `ferx_fit()`:

```{r fit, eval = FALSE}
fit <- ferx_fit(ex$model, ex$data)
fit
```

### Step 4 — Re-inspect from the fitted result

After fitting you can call `ferx_model_inspect()` directly on the `ferx_fit`
object returned by `ferx_fit()` — no need to supply the model path again:

```{r inspect-post, eval = FALSE}
ferx_model_inspect(fit)
```

The output is identical to the pre-fit inspection. This is convenient when you
want to confirm the structure inside a script or report without keeping the
model path in scope.

---

## Pipe-based workflow

`ferx_model()` bundles model and data paths into a single object that flows
through a `|>` chain. This is the preferred style when you need to adjust
options before fitting or chain several post-fit steps.

### Entry point

```{r ferx-model}
ex <- ferx_example("warfarin")
m <- ferx_model(ex$data, ex$model)
m   # prints path, data path, and structure summary
```

### Minimal pipe — inspect, fit, summarise

`ferx_fit()` dispatches on the `ferx_model` object and picks up `$data`
automatically.

```{r minimal-pipe, eval = FALSE}
ferx_model(ex$data, ex$model) |>
  ferx_fit(method = "focei", covariance = TRUE) |>
  summary()
```

### Peek at a section mid-pipe

`ferx_get_section()` prints the named section to the console and passes the
`ferx_model` object through unchanged, so the pipe continues:

```{r get-section}
ferx_model(ex$data, ex$model) |>
  ferx_get_section("parameters")
```

```{r get-section-fit, eval = FALSE}
# Print [parameters] then fit without interrupting the chain
fit <- ferx_model(ex$data, ex$model) |>
  ferx_get_section("parameters") |>
  ferx_fit(method = "focei")
```

### Override fit options before fitting

`ferx_set_section()` rewrites a section on disk and returns the `ferx_model`
unchanged.
**Important:** it edits the file in place — copy first if you do not want to
modify the original.

```{r set-section, eval = FALSE}
model_copy <- file.path(tempdir(), "warfarin.ferx")
file.copy(ex$model, model_copy)

fit <- ferx_model(ex$data, model_copy) |>
  ferx_set_section("fit_options", c(
    "  method     = focei",
    "  maxiter    = 500",
    "  covariance = true"
  )) |>
  ferx_fit()

summary(fit)
ferx_model_inspect(fit)    # no path needed after fitting
ferx_cor_matrix(fit)       # parameter correlation matrix
ferx_plot_trace(fit)       # OFV + gradient norm convergence plot
```

### Validate before a long run

`ferx_check_init()` runs only 5 iterations and returns a trace and a summary
table. Use it to confirm the OFV is dropping before committing to a long
estimation:

```{r check-init, eval = FALSE}
chk <- ferx_check_init(ex$model, ex$data, method = "focei")
chk$summary      # ofv_start, ofv_end, ofv_drop — is OFV decreasing?
ferx_plot_trace(chk$fit)   # visual check: first few iterations
```

If `ofv_drop` is small or negative, the initial values are likely poor — adjust
them in `[parameters]` or widen the parameter bounds before proceeding.

### Multi-stage method chain

Pass a character vector to `method` to chain stages — each stage starts from
the previous stage's converged parameters:

```{r method-chain, eval = FALSE}
fit <- ferx_model(ex$data, ex$model) |>
  ferx_fit(method = c("saem", "focei"), covariance = TRUE)
```

### Simulate and predict from fitted parameters

```{r sim-pred, eval = FALSE}
# Simulate 100 replicates using individual parameters from the fit
sim <- ferx_simulate(ex$model, ex$data, n_sim = 100, seed = 42, fit = fit)

# Population predictions using fitted theta (eta = 0)
pred <- ferx_predict(ex$model, ex$data, fit = fit)
```

### Post-fit diagnostics

All diagnostic outputs are fields on the `ferx_fit` object:

```{r diagnostics, eval = FALSE}
head(fit$sdtab)                # PRED, IPRED, CWRES, IWRES per observation
head(fit$ebe_etas)             # empirical Bayes ETAs per subject
head(fit$individual_estimates) # individual PK parameters (CL, V, KA, ...)
fit$eta_normality              # Shapiro-Wilk normality test per ETA
```

`ferx_estimates()` collects all estimated parameters (theta, omega, sigma) into
a tidy data frame with SEs, %RSE, and 95 % CIs. Run it after any fit that used
`covariance = TRUE`. The `transform` column reflects the parameterisation
reported by the backend (`"identity"`, `"log"`, `"logit"`, or `"variance"` for
omega rows):

```{r estimates, eval = FALSE}
ferx_estimates(fit)
#   param           transform  estimate     se  rse_pct  lower_95  upper_95  ...
#   TVCL            identity     0.134  0.0108     8.1     0.113     0.155
#   TVV             identity     7.94   0.420      5.3     7.12      8.76
#   TVKA            identity     1.07   0.160     14.9     0.756     1.38
#   OMEGA(ETA_CL)   variance     0.042  0.012     28.6    0.019     0.066
#   OMEGA(ETA_V)    variance     0.020  0.009     44.9    0.002     0.037
#   OMEGA(ETA_KA)   variance     0.157  0.048     30.6    0.063     0.251
#   PROP_ERR        proportional 0.011  0.001      9.1    0.009     0.013
```

`ferx_cor_matrix()` converts the parameter covariance matrix to correlations.
A value near ±1 between two parameters signals a structural identifiability
problem worth investigating:

```{r cor-matrix, eval = FALSE}
ferx_cor_matrix(fit)
```

**Goodness-of-fit plots** from `fit$sdtab` with ggplot2:

```{r gof-plots, eval = FALSE}
library(ggplot2)

sdtab <- fit$sdtab

# DV vs PRED
ggplot(sdtab, aes(PRED, DV)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, colour = "steelblue") +
  labs(title = "DV vs PRED", x = "Population prediction", y = "Observed")

# DV vs IPRED
ggplot(sdtab, aes(IPRED, DV)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, colour = "steelblue") +
  labs(title = "DV vs IPRED", x = "Individual prediction", y = "Observed")

# CWRES vs TIME
ggplot(sdtab, aes(TIME, CWRES)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = c(-2, 0, 2), linetype = c("dashed", "solid", "dashed"),
             colour = "steelblue") +
  labs(title = "CWRES vs TIME")

# CWRES vs PRED
ggplot(sdtab, aes(PRED, CWRES)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = c(-2, 0, 2), linetype = c("dashed", "solid", "dashed"),
             colour = "steelblue") +
  labs(title = "CWRES vs PRED")
```

ETA-covariate correlations — identifies covariates worth testing in a formal
covariate search. Pass the original data frame (not the path); any numeric
column that is constant within a subject and not a standard NONMEM column
(TIME, DV, AMT, EVID, MDV, CMT, RATE) is treated as a covariate:

```{r eta-cov, eval = FALSE}
# The warfarin dataset has no covariate columns. The following example uses
# two_cpt_oral_cov, which has body weight (WT) and creatinine clearance (CRCL).
ex_cov <- ferx_example("two_cpt_oral_cov")
fit_cov <- ferx_fit(ex_cov$model, ex_cov$data)
data_cov <- read.csv(ex_cov$data)
ferx_eta_cov(fit_cov, data_cov)
#   eta      covariate    r   p_val  flag
#   ETA_CL   WT        0.38  0.008   [!]
#   ETA_CL   CRCL      0.41  0.003   [!]
#   ETA_V1   WT        0.29  0.047
# [!] flags |r| >= 0.3; flag covariates worth formal testing
```

---

## Parameter transforms

ferx supports log and logit transforms for both thetas and ETAs directly in the
model DSL. When a transform is in use, `print()` and `ferx_estimates()` display
results on the natural scale alongside the estimation-scale values.

```r
# Log-normal CL (typical in PK — ETA_CL is additive on the log scale)
CL = TVCL * exp(ETA_CL)  # TVCL is on the natural scale; BSV is log-normal

# Logit-transformed bioavailability (constrained to (0, 1))
F = inv_logit(logit(THETA_F) + ETA_F)
```

See `vignette("parameter-transforms")` for worked examples with bundled models.

---

## Saving and reloading fits

Long estimation runs can be saved to disk and reloaded in a later session with
`ferx_save_fit()` / `ferx_load_fit()`. The saved file is a compressed RDS and
preserves the full `ferx_fit` object including `sdtab`, `ebe_etas`, trace path,
and all metadata.

```{r save-load, eval = FALSE}
# Save after a long run
fit <- ferx_fit(ex$model, ex$data, method = "focei", covariance = TRUE)
ferx_save_fit(fit, "warfarin_focei.rds")

# Later — restore and continue working
fit2 <- ferx_load_fit("warfarin_focei.rds")
identical(fit$theta, fit2$theta)   # TRUE
ferx_estimates(fit2)
```

The loaded object is indistinguishable from the original: all downstream
functions (`ferx_estimates()`, `ferx_cor_matrix()`, `ferx_plot_trace()`,
`ferx_simulate()`, etc.) work without modification.

**Tip:** save immediately after fitting so a crash or session restart does not
lose results. Use a naming convention that records the run number, method, and
date:

```{r save-naming, eval = FALSE}
ferx_save_fit(fit, sprintf("run%02d_focei_%s.rds", run_number, Sys.Date()))
```

---

## Summary

### Path-based

```r
ex <- ferx_example("warfarin")
ferx_model_inspect(ex$model)          # verify structure before fitting
fit <- ferx_fit(ex$model, ex$data)
ferx_model_inspect(fit)               # re-inspect post-fit (no path needed)
```

### Pipe-based

```r
ex <- ferx_example("warfarin")

# Inspect, fit, summarise in one chain
ferx_model(ex$data, ex$model) |>
  ferx_get_section("parameters") |>  # optional: peek before fitting
  ferx_fit(method = "focei", covariance = TRUE) |>
  summary()

# Edit options then fit (copy model file first)
model_copy <- file.path(tempdir(), "warfarin.ferx")
file.copy(ex$model, model_copy)

ferx_model(model_copy, data = ex$data) |>
  ferx_set_section("fit_options", c(
    "  method     = focei",
    "  maxiter    = 500",
    "  covariance = true"
  )) |>
  ferx_fit() |>
  summary()
```
