---
title: "Model workflow: inspect, edit, fit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model workflow: inspect, edit, fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ferx)
```

## Overview

This vignette covers two complementary workflows:

1. **Path-based workflow** — pass file paths directly to `ferx_fit()`. Straightforward for one-off fits.
2. **Pipe-based workflow** — bundle paths into a `ferx_model` object and chain steps with `|>`. Cleaner when you need to tweak options before fitting or chain post-fit steps.

Both workflows produce the same `ferx_fit` result.

**Entry point:** `ferx_model(data, model)` takes the data file first, which lets a data object flow naturally into a pipeline:

```r
ex$data |> ferx_model(ex$model) |> ferx_fit() |> summary()
```

You can also call it positionally or by name:

```r
ferx_model(ex$data, ex$model) |> ferx_fit() |> summary()
ferx_model(data = ex$data, model = ex$model) |> ferx_fit() |> summary()
```

**Migration note:** Earlier versions used the order `ferx_model(model, data)`. Old-style positional calls (`ferx_model("pk.ferx")` or `ferx_model("pk.ferx", "data.csv")`) are detected by the `.ferx` extension and auto-corrected with a deprecation warning; this shim will be removed in a future release. Calls that pass `data` by name (`ferx_model("pk.ferx", data = "data.csv")`) keep working unchanged.

---

## Path-based workflow

### Step 1 — Get a model file

`ferx_example()` returns paths to the bundled example models and their data:

```{r example}
ex <- ferx_example("warfarin")
ex$model  # path to warfarin.ferx
ex$data   # path to warfarin.csv
```

In a real workflow you would supply your own `.ferx` file here.

### Step 2 — Inspect before fitting

`ferx_model_inspect()` parses the model file and prints a concise summary of
its structure without running the optimizer. Use this to verify that ferx
interprets your model the way you expect before committing to a long run.

```{r inspect-pre}
ferx_model_inspect(ex$model)
```

The printed summary has four lines:

| Line | Meaning |
|------|---------|
| **Structural** | Detected PK model type and population parameter (theta) names |
| **IIV** | Inter-individual variability (omega) parameter names |
| **IOV** | Inter-occasion variability (kappa) parameter names, or `none` |
| **Residual** | Error model type (`proportional`, `additive`, `combined`, or `additive (log-transformed)` for log-transform-both-sides) |

> **Log-transform-both-sides (LTBS).** Writing the error model as
> `log(DV) ~ additive(SIG)` (DV on the natural scale; the engine log-transforms
> it) or `DV ~ log_additive(SIG)` (DV already log-transformed in the data) fits
> on the log scale with additive error — the equivalent of NONMEM's
> `Y = LOG(F) + EPS(1)`. Under LTBS the fit output is reported on the **log
> scale**: `IPRED`, `PRED`, `CWRES`, `IWRES` (in `sdtab`) and `DV_SIM` (from
> `ferx_simulate()`) are all log-scale, so apply `exp()` before plotting on the
> original concentration scale.

`ferx_model_inspect()` returns the structure as a named list (invisibly), which
is useful for programmatic checks:

```{r inspect-programmatic}
s <- ferx_model_inspect(ex$model)
s$theta_names   # population parameter names
s$model_type    # short label for the PK model
s$iiv           # IIV parameter names
s$residual      # error model type
```

If the printed summary does not match your expectations — for example the model
type shows `NULL` or a parameter name is missing — fix the `.ferx` file before
proceeding. This catches typos and structural mistakes cheaply, before the
optimizer runs.

### Step 3 — Fit

Once the structure looks correct, pass the same paths to `ferx_fit()`:

```{r fit, eval = FALSE}
fit <- ferx_fit(ex$model, ex$data)
fit
```

### Step 4 — Re-inspect from the fitted result

After fitting you can call `ferx_model_inspect()` directly on the `ferx_fit`
object returned by `ferx_fit()` — no need to supply the model path again:

```{r inspect-post, eval = FALSE}
ferx_model_inspect(fit)
```

The output is identical to the pre-fit inspection. This is convenient when you
want to confirm the structure inside a script or report without keeping the
model path in scope.

---

## Pipe-based workflow

`ferx_model()` bundles model and data paths into a single object that flows
through a `|>` chain. This is the preferred style when you need to adjust
options before fitting or chain several post-fit steps.

### Entry point

```{r ferx-model}
ex <- ferx_example("warfarin")
m <- ferx_model(ex$data, ex$model)
m   # prints path, data path, and structure summary
```

### Minimal pipe — inspect, fit, summarise

`ferx_fit()` dispatches on the `ferx_model` object and picks up `$data`
automatically.

```{r minimal-pipe, eval = FALSE}
ferx_model(ex$data, ex$model) |>
  ferx_fit(method = "focei", covariance = TRUE) |>
  summary()
```

### Get a section

`ferx_model_get_section()` returns the lines of a named section (accepting
either a `ferx_model` object or a plain path) and prints them to the console:

```{r get-section}
ferx_model_get_section(ex$model, "parameters")
```

### Override fit options before fitting

`ferx_model_set_section()` rewrites a section on disk and returns `x` so the
pipe can continue.

**Copy-on-write for bundled examples:** when the model is inside the installed
ferx package directory (i.e. came from `ferx_example()`), `ferx_model_set_section()`
transparently copies it to `tempdir()` first and edits the copy; the bundled
file is never modified. For user-owned files it edits in place. Copy the file
yourself if you want a stable user-controlled path.

```{r set-section, eval = FALSE}
fit <- ferx_model(ex$data, ex$model) |>
  ferx_model_set_section("fit_options", c(
    "  method     = focei",
    "  maxiter    = 500",
    "  covariance = true"
  )) |>
  ferx_fit()

summary(fit)
ferx_model_inspect(fit)    # no path needed after fitting
fit$cor_matrix             # parameter correlation matrix
plot(fit)                  # OFV + gradient norm convergence plot
```

### Validate before a long run

`ferx_check_init()` runs only 5 iterations and returns a trace and a summary
table. Use it to confirm the OFV is dropping before committing to a long
estimation:

```{r check-init, eval = FALSE}
chk <- ferx_check_init(ex$model, ex$data, method = "focei")
chk$summary      # ofv_start, ofv_end, ofv_drop — is OFV decreasing?
plot(chk$fit)              # visual check: first few iterations
```

If `ofv_drop` is small or negative, the initial values are likely poor — adjust
them in `[parameters]` or widen the parameter bounds before proceeding.

### Multi-stage method chain

Pass a character vector to `method` to chain stages — each stage starts from
the previous stage's converged parameters:

```{r method-chain, eval = FALSE}
fit <- ferx_model(ex$data, ex$model) |>
  ferx_fit(method = c("saem", "focei"), covariance = TRUE)
```

### Simulate and predict from fitted parameters

```{r sim-pred, eval = FALSE}
# Simulate 100 replicates using individual parameters from the fit
sim <- ferx_simulate(ex$model, ex$data, n_sim = 100, seed = 42, fit = fit)

# Population predictions using fitted theta (eta = 0)
pred <- ferx_predict(ex$model, ex$data, fit = fit)
```

### Post-fit diagnostics

All diagnostic outputs are fields on the `ferx_fit` object:

```{r diagnostics, eval = FALSE}
head(fit$sdtab)                # PRED, IPRED, CWRES, IWRES per observation
head(fit$ebe_etas)             # empirical Bayes ETAs per subject
head(fit$individual_estimates) # individual PK parameters (CL, V, KA, ...)
fit$eta_normality              # Shapiro-Wilk normality test per ETA
```

`fit$estimates` collects all estimated parameters (theta, omega diagonal,
sigma, kappa diagonal if IOV is present) into a tidy data frame, populated
automatically at the end of every `ferx_fit()` call. When the fit used
`covariance = TRUE` the `se`, `rse_pct`, `lower_95`, and `upper_95` columns are
populated; without the covariance step those columns are `NA`. The `transform`
column reflects how the parameter is reported by the backend -
`"identity"`, `"log"`, `"logit"`, or `"proportional"`/`"additive"`
for sigma rows. Omega rows always use `transform = "variance"` regardless of
the ETA parameterisation (the interpretation is conveyed by `print.ferx_fit()`,
not by the tidy table). The `init_as_sd` column is `TRUE` for omega/sigma/kappa
rows where the initial value was annotated as a standard deviation in the
`.ferx` file.

The parameter labels are the bare declared names (`ETA_CL`, `PROP_ERR`);
unnamed elements fall back to `OMEGA(i,i)` / `SIGMA(i)` / `KAPPA{i}`.

```{r estimates, eval = FALSE}
fit$estimates
#     param    transform   estimate         se  rse_pct  lower_95  upper_95  init_as_sd
#      TVCL     identity 0.13273892 0.01446172  10.8949  0.104394  0.161084       FALSE
#       TVV     identity 7.69480857 0.29200479   3.7948  7.122479  8.267138       FALSE
#      TVKA     identity 0.75856259 0.03505402   4.6211  0.689857  0.827268       FALSE
#    ETA_CL     variance 0.02854340 0.00637992  22.3516  0.016039  0.041048       FALSE
#     ETA_V     variance 0.00961159 0.00216463  22.5210  0.005369  0.013854       FALSE
#    ETA_KA     variance 0.34236514 0.07687305  22.4535  0.191694  0.493036       FALSE
#  PROP_ERR proportional 0.01063348 0.00078565   7.3884  0.009094  0.012173        TRUE
# estimate_natural / lower_95_natural / upper_95_natural columns are present
# but only populated for log- or logit-transformed thetas.
```

The output above is captured from a real `gn` fit of the bundled warfarin
example with `covariance = TRUE`.

`fit$cor_matrix` converts the parameter covariance matrix to correlations.
A value near ±1 between two parameters signals a structural identifiability
problem worth investigating:

```{r cor-matrix, eval = FALSE}
fit$cor_matrix
```

**Goodness-of-fit plots** from `fit$sdtab` with ggplot2:

```{r gof-plots, eval = FALSE}
library(ggplot2)

sdtab <- fit$sdtab

# DV vs PRED
ggplot(sdtab, aes(PRED, DV)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, colour = "steelblue") +
  labs(title = "DV vs PRED", x = "Population prediction", y = "Observed")

# DV vs IPRED
ggplot(sdtab, aes(IPRED, DV)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, colour = "steelblue") +
  labs(title = "DV vs IPRED", x = "Individual prediction", y = "Observed")

# CWRES vs TIME
ggplot(sdtab, aes(TIME, CWRES)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = c(-2, 0, 2), linetype = c("dashed", "solid", "dashed"),
             colour = "steelblue") +
  labs(title = "CWRES vs TIME")

# CWRES vs PRED
ggplot(sdtab, aes(PRED, CWRES)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = c(-2, 0, 2), linetype = c("dashed", "solid", "dashed"),
             colour = "steelblue") +
  labs(title = "CWRES vs PRED")
```

ETA-covariate correlations — `fit$eta_cov` identifies covariates worth testing
in a formal covariate search. It is computed automatically from the dataset
used to fit the model; any numeric column that is constant within a subject
and not a standard NONMEM column (TIME, DV, AMT, EVID, MDV, CMT, RATE) is
treated as a covariate:

```{r eta-cov, eval = FALSE}
# The warfarin dataset has no covariate columns. The example below uses
# two_cpt_oral_cov, which carries body weight (WT) and creatinine clearance
# (CRCL). Rows are sorted by descending |r|; the `flag` column is set to
# "[!]" whenever |r| >= 0.3 to highlight pairs worth formal testing.
ex_cov  <- ferx_example("two_cpt_oral_cov")
fit_cov <- ferx_fit(ex_cov$model, ex_cov$data, method = "gn", covariance = FALSE)
fit_cov$eta_cov
#       eta covariate      r  p_val flag
#    ETA_KA        WT  0.144 0.4472
#    ETA_CL      CRCL -0.117 0.5386
#     ETA_Q        WT  0.105 0.5802
#    ETA_V1        WT -0.093 0.6251
#    ETA_V2      CRCL  0.081 0.6699
#    ...
# None reach |r| >= 0.3 in this dataset, so no rows are flagged. A real
# covariate signal would show up at the top of the table with [!] in `flag`.
```

---

## Parameter transforms

ferx supports log and logit transforms for both thetas and ETAs directly in the
model DSL. When a transform is in use, `print()` and `fit$estimates` display
results on the natural scale alongside the estimation-scale values.

```r
# Log-normal CL (typical in PK — ETA_CL is additive on the log scale)
CL = TVCL * exp(ETA_CL)  # TVCL is on the natural scale; BSV is log-normal

# Logit-transformed bioavailability (constrained to (0, 1))
F = inv_logit(logit(THETA_F) + ETA_F)
```

See `vignette("parameter-transforms")` for worked examples with bundled models.

---

## Saving and reloading fits

Long estimation runs can be saved to disk and reloaded in a later session with
`ferx_save_fit()` / `ferx_load_fit()`. The saved file is a compressed RDS and
preserves the full `ferx_fit` object including `sdtab`, `ebe_etas`, trace path,
and all metadata.

```{r save-load, eval = FALSE}
# Save after a long run
fit <- ferx_fit(ex$model, ex$data, method = "focei", covariance = TRUE)
ferx_save_fit(fit, "warfarin_focei.rds")

# Later — restore and continue working
fit2 <- ferx_load_fit("warfarin_focei.rds")
identical(fit$theta, fit2$theta)   # TRUE
fit2$estimates
```

The loaded object is indistinguishable from the original: all downstream
fields (`estimates`, `cor_matrix`, `eta_cov`) and functions
(`plot()`, `ferx_simulate()`, etc.) work without modification.

**Tip:** save immediately after fitting so a crash or session restart does not
lose results. Use a naming convention that records the run number, method, and
date:

```{r save-naming, eval = FALSE}
ferx_save_fit(fit, sprintf("run%02d_focei_%s.rds", run_number, Sys.Date()))
```

---

## Summary

### Path-based

```r
ex <- ferx_example("warfarin")
ferx_model_inspect(ex$model)          # verify structure before fitting
fit <- ferx_fit(ex$model, ex$data)
ferx_model_inspect(fit)               # re-inspect post-fit (no path needed)
```

### Pipe-based

```r
ex <- ferx_example("warfarin")

ferx_model_get_section(ex$model, "parameters")  # optional: inspect before fitting

# Inspect, fit, summarise in one chain
ferx_model(ex$data, ex$model) |>
  ferx_fit(method = "focei", covariance = TRUE) |>
  summary()

# Edit options then fit. ferx_model_set_section() auto-copies bundled examples
# to tempdir() before editing, so the package file stays pristine.
ferx_model(ex$data, ex$model) |>
  ferx_model_set_section("fit_options", c(
    "  method     = focei",
    "  maxiter    = 500",
    "  covariance = true"
  )) |>
  ferx_fit() |>
  summary()
```
