---
title: "Model workflow: inspect, edit, fit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model workflow: inspect, edit, fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ferx)
```

## Overview

This vignette covers the recommended pre-fit inspection workflow:

1. Point ferx at a model file.
2. **Inspect** the parsed structure before committing to a (potentially long) estimation run.
3. Fit the model.
4. Re-inspect the fitted result without re-supplying the file path.

## Step 1 — Get a model file

`ferx_example()` returns paths to the bundled example models and their data:

```{r example}
ex <- ferx_example("warfarin")
ex$model  # path to warfarin.ferx
ex$data   # path to warfarin.csv
```

In a real workflow you would supply your own `.ferx` file here.

## Step 2 — Inspect before fitting

`ferx_model_inspect()` parses the model file and prints a concise summary of
its structure without running the optimizer. Use this to verify that ferx
interprets your model the way you expect before committing to a long run.

```{r inspect-pre}
ferx_model_inspect(ex$model)
```

The printed summary has four lines:

| Line | Meaning |
|------|---------|
| **Structural** | Detected PK model type and population parameter (theta) names |
| **IIV** | Inter-individual variability (omega) parameter names |
| **IOV** | Inter-occasion variability (kappa) parameter names, or `none` |
| **Residual** | Error model type (`proportional`, `additive`, or `combined`) |

`ferx_model_inspect()` returns the structure as a named list (invisibly), which
is useful for programmatic checks:

```{r inspect-programmatic}
s <- ferx_model_inspect(ex$model)
s$theta_names   # population parameter names
s$model_type    # short label for the PK model
s$iiv           # IIV parameter names
s$residual      # error model type
```

If the printed summary does not match your expectations — for example the model
type shows `NULL` or a parameter name is missing — fix the `.ferx` file before
proceeding. This catches typos and structural mistakes cheaply, before the
optimizer runs.

## Step 3 — Fit

Once the structure looks correct, pass the same paths to `ferx_fit()`:

```{r fit, eval = FALSE}
fit <- ferx_fit(ex$model, ex$data)
fit
```

## Step 4 — Re-inspect from the fitted result

After fitting you can call `ferx_model_inspect()` directly on the `ferx_fit`
object returned by `ferx_fit()` — no need to supply the model path again:

```{r inspect-post, eval = FALSE}
ferx_model_inspect(fit)
```

The output is identical to the pre-fit inspection. This is convenient when you
want to confirm the structure inside a script or report without keeping the
model path in scope.

## Summary

```r
# 1. Get model and data paths
ex <- ferx_example("warfarin")

# 2. Inspect before fitting — verify structure without running the optimizer
ferx_model_inspect(ex$model)

# 3. Fit
fit <- ferx_fit(ex$model, ex$data)

# 4. Re-inspect post-fit (no path required)
ferx_model_inspect(fit)
```
