---
title: "Parameter transforms: log, logit, and additive ETAs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameter transforms: log, logit, and additive ETAs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ferx)
```

## Overview

ferx supports three parameterisation styles for individual parameters:

| Style | DSL pattern | Use case |
|---|---|---|
| **Log-normal** | `P = TVP * exp(ETA_P)` | Most PK parameters (CL, V, KA) — positive by construction |
| **Logit-normal** | `P = inv_logit(logit(THETA_P) + ETA_P)` | Fractions/probabilities (F, bioavailability) — bounded (0, 1) |
| **Additive** | `P = TVP + ETA_P` | Parameters with natural additive variability (lag time, threshold) |

The optimiser always works on the unbounded or transformed scale; ferx
back-transforms to the natural scale when reporting results.

---

## Log-normal ETAs (default)

The standard warfarin model uses log-normal ETAs for CL, V, and KA. This is
the default: each individual parameter equals the typical value multiplied by
`exp(ETA)`, which keeps the parameter positive and gives a log-normal
distribution across subjects.

```{r logn-inspect}
ex <- ferx_example("warfarin")
ferx_model_inspect(ex$model)
```

After fitting, `print()` reports omega as `[log-normal]` with a CV%:

```{r logn-fit, eval = FALSE}
fit <- ferx_fit(ex$model, ex$data)
fit   # omega block shows "[log-normal]  CV% = ..."
```

`ferx_estimates()` returns the omega variance on the log scale in `estimate`,
with `transform = "log_normal"`:

```{r logn-estimates, eval = FALSE}
ferx_estimates(fit)
```

---

## Logit-normal ETA (bioavailability)

When a parameter is a fraction bounded between 0 and 1, use `inv_logit` /
`logit` to ensure the individual values stay in (0, 1):

```
F = inv_logit(logit(THETA_F) + ETA_F)
```

`THETA_F` is specified on the natural (0, 1) scale in `[parameters]`; ferx
maps it to the logit scale internally so the optimiser works on an unbounded
space. The omega for `ETA_F` is the variance on the logit scale.

```{r logit-inspect}
ex_logit <- ferx_example("warfarin_logit_f")
ferx_model_inspect(ex_logit$model)
```

After fitting, `print()` shows:

- `THETA_F` with `[logit scale]` and a 95 % CI back-transformed through
  `inv_logit` so the CI is on the (0, 1) scale.
- `ETA_F` with `[logit]`, `SD_logit`, and a ±1 SD interval on the (0, 1)
  scale.

```{r logit-fit, eval = FALSE}
fit_logit <- ferx_fit(ex_logit$model, ex_logit$data)
fit_logit
```

`ferx_estimates()` populates `estimate_natural`, `lower_95_natural`, and
`upper_95_natural` for logit-transformed thetas:

```{r logit-estimates, eval = FALSE}
est <- ferx_estimates(fit_logit)
est[est$param == "THETA_F", ]
# estimate         — logit scale
# estimate_natural — (0, 1) scale (inv_logit back-transform)
# lower_95_natural / upper_95_natural — asymmetric CI on (0, 1) scale
```

---

## Additive ETA (lag time)

When the BSV on a parameter has an additive rather than multiplicative
structure — for example, a lag time where the subject-level deviation is in
hours — use a plain sum:

```
TLAG = TVTLAG + ETA_TLAG
```

`ETA_TLAG` follows a normal distribution with mean 0 and variance given by the
corresponding omega element. The individual lag times are therefore normally
distributed around `TVTLAG`.

```{r add-inspect}
ex_add <- ferx_example("warfarin_additive_eta")
ferx_model_inspect(ex_add$model)
```

After fitting, the `ETA_TLAG` row in the omega block is printed with
`[additive]` and an SD in the original units (hours):

```{r add-fit, eval = FALSE}
fit_add <- ferx_fit(ex_add$model, ex_add$data)
fit_add
```

`ferx_estimates()` returns `transform = "additive"` for that omega element;
the CI is symmetric (no back-transform needed):

```{r add-estimates, eval = FALSE}
ferx_estimates(fit_add)
```

---

## Mixing transforms in one model

All three styles can appear in the same model. The standard two-compartment
oral model with a logit bioavailability and an additive lag time would declare:

```
[individual_parameters]
  CL   = TVCL  * exp(ETA_CL)        # log-normal
  V1   = TVV1  * exp(ETA_V1)        # log-normal
  KA   = TVKA  * exp(ETA_KA)        # log-normal
  F    = inv_logit(logit(THETA_F) + ETA_F)   # logit-normal
  TLAG = TVTLAG + ETA_TLAG          # additive
```

Each ETA is handled independently; the omega matrix contains the corresponding
variances on the respective scales (log, logit, or natural units).

---

## Summary table

| Transform | Omega interpretation | Print label | `ferx_estimates()` columns |
|---|---|---|---|
| Log-normal | variance on log scale | `[log-normal]` + CV% | `estimate` (log var), `transform = "log_normal"` |
| Logit-normal | variance on logit scale | `[logit]` + SD_logit + ±1 SD interval | same; theta gets `estimate_natural` etc. |
| Additive | variance in original units | `[additive]` + SD | `estimate` (variance), symmetric CI |
