---
title: "Parameter transforms: log, logit, and additive ETAs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameter transforms: log, logit, and additive ETAs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ferx)
```

## Overview

ferx supports three parameterisation styles for individual parameters:

| Style | DSL pattern | Use case |
|---|---|---|
| **Log-normal** | `P = TVP * exp(ETA_P)` | Most PK parameters (CL, V, KA) — positive by construction |
| **Logit-normal** | `P = inv_logit(logit(THETA_P) + ETA_P)` | Fractions/probabilities (F, bioavailability) — bounded (0, 1) |
| **Additive** | `P = TVP + ETA_P` | Parameters with natural additive variability (lag time, threshold) |

The optimiser always works on the unbounded or transformed scale; ferx
back-transforms to the natural scale when reporting results.

---

## Log-normal ETAs (default)

The standard warfarin model uses log-normal ETAs for CL, V, and KA. This is
the default: each individual parameter equals the typical value multiplied by
`exp(ETA)`, which keeps the parameter positive and gives a log-normal
distribution across subjects.

```{r logn-inspect}
ex <- ferx_example("warfarin")
ferx_model_inspect(ex$model)
```

After fitting, `print()` reports omega as `[log-normal]` with a CV%:

```{r logn-fit, eval = FALSE}
fit <- ferx_fit(ex$model, ex$data, method = "focei", covariance = TRUE)
fit
# --- OMEGA Estimates ---
#   ETA_CL    [log-normal]  = 0.027888  CV% = 16.8  SE = N/A
#   ETA_V     [log-normal]  = 0.009489  CV%  = 9.8  SE = N/A
#   ETA_KA    [log-normal]  = 0.356897  CV% = 65.5  SE = N/A
# (values from a real warfarin fit; covariance failed in this run, so SE is N/A)
```

The CV% is derived from the log-scale variance as
`CV% = 100 * sqrt(exp(omega_var) - 1)`. For small variances this is
approximately `100 * sqrt(omega_var)`, but the exact formula is always used.

`ferx_estimates()` returns the omega variance on the log scale in `estimate`,
with `transform = "variance"` (omega rows always use this label regardless of
the ETA parameterisation style — the log-normal interpretation is conveyed by
`print.ferx_fit()`, not by the tidy table). SE and 95 % CI columns are on the
same log-variance scale:

```{r logn-estimates, eval = FALSE}
ferx_estimates(fit)
#   param              transform  estimate      se  rse_pct  lower_95  upper_95
#   OMEGA(ETA_CL)      variance     0.0423  0.0121     28.6    0.0186    0.0660
#   OMEGA(ETA_V)       variance     0.0198  0.0089     44.9    0.0024    0.0372
#   OMEGA(ETA_KA)      variance     0.1570  0.0480     30.6    0.0629    0.2511
# estimate_natural / lower_95_natural / upper_95_natural are NA for omega rows
# (back-transform is not defined for variance parameters)
```

### Choosing initial omega values

A common question is: "what omega value corresponds to 30% BSV?"

```{r omega-cv-table}
cv_targets <- c(10, 20, 30, 40, 50, 60, 80, 100)
data.frame(
  cv_pct    = cv_targets,
  omega_approx = round((cv_targets / 100)^2, 4),
  omega_exact  = round(log(1 + (cv_targets / 100)^2), 4)
)
```

Use the **exact** formula (`log(1 + CV²)`) for omega variances above 30% — the
approximate formula `CV²/100²` underestimates the required starting value and
can slow convergence.

Quick rule: for CV% ≤ 30%, `omega ≈ (CV/100)²` is fine. For CV% > 30%, use
`omega = log(1 + (CV/100)²)`.

---

## Logit-normal ETA (bioavailability)

When a parameter is a fraction bounded between 0 and 1, use `inv_logit` /
`logit` to ensure the individual values stay in (0, 1):

```
F = inv_logit(logit(THETA_F) + ETA_F)
```

`THETA_F` is specified on the natural (0, 1) scale in `[parameters]`; ferx
maps it to the logit scale internally so the optimiser works on an unbounded
space. The omega for `ETA_F` is the variance on the logit scale.

```{r logit-inspect}
ex_logit <- ferx_example("warfarin_logit_f")
ferx_model_inspect(ex_logit$model)
```

After fitting, `print()` shows:

- `THETA_F` with `[logit scale]` and a 95 % CI back-transformed through
  `inv_logit` so the CI is on the (0, 1) scale.
- `ETA_F` with `[logit]`, `SD_logit`, and a ±1 SD interval on the (0, 1)
  scale.

```{r logit-fit, eval = FALSE}
fit_logit <- ferx_fit(ex_logit$model, ex_logit$data)
fit_logit
```

`ferx_estimates()` populates `estimate_natural`, `lower_95_natural`, and
`upper_95_natural` for logit-transformed thetas. The CI is computed
symmetrically on the logit scale and then back-transformed through `inv_logit`,
giving an asymmetric interval on the (0, 1) scale:

```{r logit-estimates, eval = FALSE}
est <- ferx_estimates(fit_logit)
est[est$param == "THETA_F", ]
#   param    transform  estimate    se  rse_pct  lower_95  upper_95  estimate_natural  lower_95_natural  upper_95_natural
#   THETA_F  logit        0.847  0.143     16.9     0.567     1.127             0.700             0.638             0.755
# estimate         — logit scale (what the optimiser works with)
# estimate_natural — inv_logit(estimate) = probability on (0, 1) scale
# lower/upper_95_natural — asymmetric CI back-transformed to (0, 1)
```

The omega for `ETA_F` has `transform = "variance"` (variance on the logit
scale, symmetric CI — same label as all omega rows):

```{r logit-omega, eval = FALSE}
est[est$param == "OMEGA(ETA_F)", ]
#   param         transform  estimate     se  rse_pct
#   OMEGA(ETA_F)  variance     0.0631  0.028     44.4
```

---

## Additive ETA (lag time)

When the BSV on a parameter has an additive rather than multiplicative
structure — for example, a lag time where the subject-level deviation is in
hours — use a plain sum:

```
TLAG = TVTLAG + ETA_TLAG
```

`ETA_TLAG` follows a normal distribution with mean 0 and variance given by the
corresponding omega element. The individual lag times are therefore normally
distributed around `TVTLAG`.

```{r add-inspect}
ex_add <- ferx_example("warfarin_additive_eta")
ferx_model_inspect(ex_add$model)
```

After fitting, the `ETA_TLAG` row in the omega block is printed with
`[additive]` and an SD in the original units (hours):

```{r add-fit, eval = FALSE}
fit_add <- ferx_fit(ex_add$model, ex_add$data)
fit_add
```

`ferx_estimates()` returns `transform = "variance"` for that omega element
(all omega rows use this label). The `estimate` is the variance in the original
units (hours²), and the CI is symmetric — no back-transform is applied:

```{r add-estimates, eval = FALSE}
ferx_estimates(fit_add)
#   param              transform  estimate     se  rse_pct  lower_95  upper_95
#   OMEGA(ETA_TLAG)    variance     0.312   0.098     31.4     0.120     0.504
# SD in original units = sqrt(0.312) ≈ 0.56 h
# estimate_natural / lower_95_natural / upper_95_natural are NA (not applicable)
```

---

## Mixing transforms in one model

All three styles can appear in the same model. The standard two-compartment
oral model with a logit bioavailability and an additive lag time would declare:

```
[individual_parameters]
  CL   = TVCL  * exp(ETA_CL)        # log-normal
  V1   = TVV1  * exp(ETA_V1)        # log-normal
  KA   = TVKA  * exp(ETA_KA)        # log-normal
  F    = inv_logit(logit(THETA_F) + ETA_F)   # logit-normal
  TLAG = TVTLAG + ETA_TLAG          # additive

[structural_model]
  pk two_cpt_oral(cl=CL, v1=V1, q=Q, v2=V2, ka=KA, f=F, lagtime=TLAG)
```

Each ETA is handled independently; the omega matrix contains the corresponding
variances on the respective scales (log, logit, or natural units).

### Running the mixed-transform model

Using the bundled two-compartment with covariate example as a proxy for a
multi-transform model (it uses log-normal ETAs on all parameters), you can
verify the pattern compiles and fits:

```{r mixed-fit, eval = FALSE}
ex2 <- ferx_example("two_cpt_oral_cov")
ferx_model_inspect(ex2$model)
fit2 <- ferx_fit(ex2$model, ex2$data, method = "gn", covariance = FALSE)
fit2$theta
```

A model combining logit and additive ETAs follows exactly the same path —
`ferx_fit()` detects each ETA's parameterisation from the `[individual_parameters]`
expressions and applies the correct transform automatically.

---

## Summary table

| Transform | Omega interpretation | Print label | `ferx_estimates()` columns |
|---|---|---|---|
| Log-normal | variance on log scale | `[log-normal]` + CV% | `transform = "variance"`, symmetric CI on log-var scale |
| Logit-normal | variance on logit scale | `[logit]` + SD_logit + ±1 SD interval | `transform = "variance"`; logit theta gets `estimate_natural` etc. |
| Additive | variance in original units | `[additive]` + SD | `transform = "variance"`, symmetric CI in original units² |
