---
title: "Simulation, prediction, and uncertainty propagation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation, prediction, and uncertainty propagation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ferx)
```

## Overview

After a model has been fit, three functions cover the standard
post-estimation workflow:

| Function | Returns | Captures |
|---|---|---|
| `ferx_predict()` | population predictions (`PRED`) | typical-value trajectory (eta = 0) |
| `ferx_simulate()` | replicates of `IPRED` and `DV_SIM` | between-subject variability + residual error |
| `ferx_simulate_with_uncertainty()` | same + parameter draws | also propagates parameter uncertainty |

All three take the same `(model, data)` pair as `ferx_fit()` and accept an
optional `fit =` argument that substitutes the fitted parameter values for the
model file's initial estimates. When `fit` is omitted, the initial values from
the `.ferx` file are used - which is occasionally useful for sanity-checking a
model before fitting, but the post-fit case is the common one.

---

## Setup

```{r setup-fit, eval = FALSE}
ex  <- ferx_example("warfarin")
fit <- ferx_fit(ex$model, ex$data, method = "gn", covariance = TRUE)
```

The examples below assume `ex` and `fit` are in scope.

---

## Population prediction

`ferx_predict()` returns one row per observation in the data, with the
typical-value prediction (eta = 0) for each subject's covariates and dosing
record:

```{r predict, eval = FALSE}
pred <- ferx_predict(ex$model, ex$data, fit = fit)
head(pred)
#>   ID TIME      PRED
#> 1  1  0.5  4.083326
#> 2  1  1.0  6.842694
#> 3  1  2.0  9.930368
#> 4  1  4.0 11.771746
#> 5  1  8.0 11.553212
#> 6  1 12.0 10.810149
```

Without `fit =`, the `[parameters]` block's initial estimates are used. The
columns are always `ID, TIME, PRED`.

---

## Individual simulation

`ferx_simulate()` draws fresh etas and epsilons and returns simulated
observations for each replicate. Each row carries:

| Column | Meaning |
|---|---|
| `DRAW`  | Always `1` for `ferx_simulate()`; present so downstream code can share aggregation logic with `ferx_simulate_with_uncertainty()` |
| `SIM`   | Replicate index, 1..`n_sim` |
| `ID`    | Subject ID from the data file |
| `TIME`  | Observation time from the data file |
| `IPRED` | Individual prediction (using the drawn eta) |
| `DV_SIM`| Simulated observation (IPRED plus residual error) |

```{r simulate, eval = FALSE}
sim <- ferx_simulate(ex$model, ex$data, n_sim = 3L, seed = 1L, fit = fit)
head(sim)
#>   DRAW SIM ID TIME     IPRED    DV_SIM
#> 1    1   1  1  0.5  3.566685  3.543429
#> 2    1   1  1  1.0  6.037146  5.978719
#> 3    1   1  1  2.0  8.895675  9.077145
#> 4    1   1  1  4.0 10.700945 10.683510
#> 5    1   1  1  8.0 10.468944 10.517356
#> 6    1   1  1 12.0  9.654178  9.659524

nrow(sim)            # n_sim * n_obs
#> [1] 330
unique(sim$SIM)
#> [1] 1 2 3
```

`seed` makes the draws reproducible across runs.

### Quick visual prediction interval

A common diagnostic is a per-time prediction interval over the simulated
replicates. With `n_sim = 100` and base aggregation:

```{r vpc, eval = FALSE}
sim <- ferx_simulate(ex$model, ex$data, n_sim = 100L, seed = 1L, fit = fit)

pi <- aggregate(DV_SIM ~ TIME, data = sim,
                FUN = function(x) quantile(x, c(0.05, 0.5, 0.95)))
head(pi)
#>   TIME DV_SIM.5% DV_SIM.50% DV_SIM.95%
#> 1  0.5  1.722816   4.035638   8.355966
#> 2  1.0  3.183157   6.773509  11.451650
#> 3  2.0  5.555958   9.734206  13.054238
#> 4  4.0  8.371048  11.408090  13.834516
#> 5  8.0  9.616953  11.337663  13.268086
#> 6 12.0  9.221792  10.715382  12.470054
```

To draw a visual predictive check with ggplot2 you would unpack the matrix
column produced by `quantile()` into named columns and layer the ribbon over
the observed `DV`:

```{r vpc-plot, eval = FALSE}
library(ggplot2)

q <- as.data.frame(pi$DV_SIM)
names(q) <- c("lo", "med", "hi")
pi_df <- data.frame(TIME = pi$TIME, q)

obs <- read.csv(ex$data)
obs <- obs[obs$EVID == 0 & obs$MDV == 0, ]

ggplot() +
  geom_ribbon(data = pi_df, aes(TIME, ymin = lo, ymax = hi),
              fill = "steelblue", alpha = 0.25) +
  geom_line(data   = pi_df, aes(TIME, med),
            colour = "steelblue") +
  geom_point(data  = obs, aes(TIME, DV),
             alpha = 0.4) +
  labs(title = "Simulation-based 90% prediction interval",
       y     = "DV")
```

The simple aggregation above stratifies only by `TIME`. For a "real" VPC you
would also stratify by dose or covariate group, and bin times for sparse
designs - both are downstream data-frame manipulations and are not built into
ferx.

---

## Propagating parameter uncertainty

`ferx_simulate_with_uncertainty()` adds an outer loop: for each of
`n_uncertainty_draws` parameter sets drawn from the fit's uncertainty
distribution, `n_sim_per_draw` eta/eps replicates are simulated. The result
includes both the within-replicate variability and the parameter uncertainty
itself.

Two uncertainty sources are supported:

| `method` | Requires | Sampling |
|---|---|---|
| `"asymptotic"` (default) | `fit$cov_matrix` (i.e. `ferx_fit(..., covariance = TRUE)`) | MVN draws in the engine's packed (log-theta, Cholesky-omega, log-sigma) space |
| `"sir"` | `fit$sir_resamples` (i.e. `ferx_fit(..., sir = TRUE, settings = list(sir_keep_samples = TRUE))`) | Resampling with replacement from the stored SIR resamples |

The return value has the same column set as `ferx_simulate()`, but `DRAW` now
ranges over 1..`n_uncertainty_draws`:

```{r sim-unc, eval = FALSE}
sims <- ferx_simulate_with_uncertainty(
  ex$model, ex$data, fit,
  n_uncertainty_draws = 5L,
  n_sim_per_draw      = 2L,
  seed                = 1L
)
head(sims)
#>   DRAW SIM ID TIME     IPRED    DV_SIM
#> 1    1   1  1  0.5  4.304992  4.259951
#> 2    1   1  1  1.0  7.313132  7.408912
#> 3    1   1  1  2.0 10.827313 10.718364
#> 4    1   1  1  4.0 13.042761 13.197373
#> 5    1   1  1  8.0 12.568896 12.410489
#> 6    1   1  1 12.0 11.337537 11.548863

nrow(sims)
#> [1] 1100   # 5 draws * 2 sims/draw * 110 obs
```

The row count is `n_uncertainty_draws * n_sim_per_draw * n_obs`. For an
uncertainty-aware prediction interval, aggregate over all rows at each `TIME`:

```{r sim-unc-pi, eval = FALSE}
pi_unc <- aggregate(DV_SIM ~ TIME, data = sims,
                    FUN = function(x) quantile(x, c(0.05, 0.5, 0.95)))
head(pi_unc)
```

The asymptotic method is fast but assumes the parameter likelihood is locally
quadratic around the ML estimate. For models with strong non-normality or
boundary issues, prefer the SIR method - but only when the cost of running SIR
at fit time is acceptable.

---

## Choosing between the three

- **`ferx_predict()`** when you only need the typical-value trajectory - for a
  PRED-vs-time overlay, or for forecasting under a new dosing schedule.
- **`ferx_simulate()`** when you want a prediction interval that reflects
  between-subject variability and residual error, but you are willing to treat
  the ML estimates as fixed.
- **`ferx_simulate_with_uncertainty()`** when the question is about uncertainty
  in the parameter values themselves: dose-recommendation intervals, model
  comparison via predictive performance, or any decision-relevant interval
  that would understate spread if parameter uncertainty were ignored.
