---
title: "SIR uncertainty quantification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIR uncertainty quantification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ferx)
```

## Overview

Sampling Importance Resampling (SIR) refines the uncertainty estimate from a
fitted model without requiring a new optimisation run. It draws candidate
parameter vectors from the asymptotic proposal (the fit's covariance matrix),
re-weights each candidate by the likelihood ratio against the proposal, and
keeps a resample drawn in proportion to those weights.

In ferx, SIR is the alternative to the asymptotic (Wald) confidence intervals.
Use it when:

- the asymptotic CI may be a poor approximation (skewed likelihood, parameter
  near a boundary, small sample size);
- you want to propagate parameter uncertainty into downstream simulation via
  `ferx_simulate_with_uncertainty(method = "sir")`.

Two entry points produce the same output structure:

| Entry point | When to use |
|---|---|
| `ferx_fit(..., sir = TRUE)` | Inline at fit time. Convenient when SIR is part of the routine workflow. |
| `ferx_sir(fit, ...)` | Post-hoc against an existing fit (including one loaded via `ferx_load_fit()`). The fit must have been computed with `covariance = TRUE`. |

`ferx_sir()` re-reads the model and data files recorded on the fit, hashes
them, and **hard-errors** if either has changed since fitting - SIR against a
modified model or dataset is meaningless. The integrity check is documented
in `?ferx_sir`.

---

## Running SIR post-hoc

```{r run-sir, eval = FALSE}
ex  <- ferx_example("warfarin")
fit <- ferx_fit(ex$model, ex$data, method = "gn", covariance = TRUE)

fit <- ferx_sir(fit,
                sir_samples   = 1000,
                sir_resamples = 250,
                sir_seed      = 42)
```

The function returns the same `fit` object augmented with SIR fields. No new
optimisation occurs; only proposal draws, importance weighting, and resampling.

### Tuning knobs

| Argument | Default | Meaning |
|---|---|---|
| `sir_samples` | 1000 | Number of candidate draws from the asymptotic proposal. Higher = tighter weights at proportional cost. |
| `sir_resamples` | 250 | Number of resampled vectors retained. Must be `<= sir_samples`. |
| `sir_seed` | `NULL` | Integer seed for reproducibility. `NULL` uses the engine's built-in seed. |
| `sir_keep_samples` | `FALSE` | Retain the resampled packed parameter vectors on the fit. Required for downstream `ferx_simulate_with_uncertainty(method = "sir")`. |

---

## Reading the output

`ferx_sir()` (or `ferx_fit(..., sir = TRUE)`) populates five fields:

```{r sir-fields, eval = FALSE}
fit$sir_ess          # effective sample size
fit$sir_ci_theta     # 95% CI for each theta
fit$sir_ci_omega     # 95% CI for the omega diagonal
fit$sir_ci_sigma     # 95% CI for sigma
# When sir_keep_samples = TRUE:
fit$sir_resamples      # flat numeric vector of resampled packed parameters
fit$sir_resamples_n    # number of resamples
fit$sir_resamples_dim  # length of each packed parameter vector
```

The `sir_ci_*` matrices have `lower` and `upper` columns and use the bare
declared parameter names as row labels (`ETA_CL`, `PROP_ERR`, etc.), falling
back to `OMEGA(i,i)` / `SIGMA(i)` for unnamed elements - the same convention
as the rest of the package output.

### Worked output

Captured from a real `gn` fit of the bundled warfarin example with
`sir_samples = 1000, sir_resamples = 250, sir_seed = 42`:

```{r sir-output, eval = FALSE}
fit$sir_ess
#> [1] 4.71

fit$sir_ci_theta
#>           lower     upper
#> TVCL  0.1201408 0.1476027
#> TVV   7.3308524 8.1238270
#> TVKA  0.5709285 0.8822086

fit$sir_ci_omega
#>              lower      upper
#> ETA_CL 0.015943626 0.07997176
#> ETA_V  0.005861108 0.02324717
#> ETA_KA 0.191731831 0.75244194

fit$sir_ci_sigma
#>                lower     upper
#> PROP_ERR 0.009124243 0.0123905
```

### Interpreting ESS

The effective sample size is the most important diagnostic to look at first.
It summarises how well the asymptotic proposal matches the true likelihood
surface. ESS is bounded above by `sir_resamples`.

- **ESS close to `sir_resamples`** - the asymptotic proposal is a good match;
  the SIR CIs will be close to the asymptotic CIs and most of the resamples
  are distinct.
- **ESS much smaller than `sir_resamples`** - the proposal is a poor match.
  A few high-weight proposals dominate the resample, the SIR CIs are based
  on relatively few effective draws, and the resamples are heavily duplicated.

In the captured output above, `sir_ess = 4.71` against `sir_resamples = 250` is
extremely low - a sign that the warfarin asymptotic covariance does not
faithfully represent the true uncertainty (the dataset is small and at least
one parameter is poorly constrained). For real analyses, increase
`sir_samples`, refit with better initial values, or treat the SIR CI with
caution if ESS stays low.

---

## Inline SIR at fit time

When SIR is part of the standard workflow, set `sir = TRUE` in `ferx_fit()`
directly. The SIR tuning knobs flow through `settings`:

```{r inline-sir, eval = FALSE}
fit <- ferx_fit(
  ex$model, ex$data,
  method      = "focei",
  covariance  = TRUE,
  sir         = TRUE,
  settings    = list(
    sir_samples       = 2000L,
    sir_resamples     = 500L,
    sir_keep_samples  = TRUE
  )
)

fit$sir_ess
fit$sir_ci_theta
```

The output structure is identical to the `ferx_sir()` flow above.

---

## Feeding SIR into uncertainty-aware simulation

To propagate the SIR resamples into a simulation, retain them with
`sir_keep_samples = TRUE` and pass the fit to
`ferx_simulate_with_uncertainty()` with `method = "sir"`:

```{r sir-simulate, eval = FALSE}
fit <- ferx_sir(fit,
                sir_samples      = 2000,
                sir_resamples    = 500,
                sir_keep_samples = TRUE)

sims <- ferx_simulate_with_uncertainty(
  ex$model, ex$data, fit,
  n_uncertainty_draws = 200,
  n_sim_per_draw      = 5,
  method              = "sir"
)
head(sims)
#>   DRAW SIM ID TIME IPRED DV_SIM
#> 1    1   1  1  0.5  ...   ...
```

Each parameter draw is now a resample from the SIR posterior rather than an
MVN draw around the ML estimate. The downstream aggregation patterns
(quantile per `TIME`, ribbon plots, etc.) from `vignette("simulation")` apply
unchanged - replace `aggregate(DV_SIM ~ TIME, ...)` with the SIR-based
simulation output.

---

## When to prefer SIR over asymptotic

| Situation | Recommended method |
|---|---|
| Quick uncertainty for a well-behaved fit | asymptotic (`covariance = TRUE` only) |
| Parameters near a boundary, skewed likelihood, small N | SIR |
| Reporting CIs for decision-relevant quantities | SIR (or both, and compare) |
| Need uncertainty-aware simulation | both work; SIR via `sir_keep_samples = TRUE`, asymptotic via the covariance matrix |

The cost is SIR draws plus likelihood evaluations - typically cheaper than a
full refit but not free. Start with `sir_samples = 1000` and increase if ESS
is low.
