---
title: "Derived quantities and custom output columns"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Derived quantities and custom output columns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ferx)
```

## Overview

Two optional blocks in a `.ferx` file let you compute post-fit summary
quantities and echo existing model quantities in `fit$sdtab`:

| Block | Purpose |
|---|---|
| `[derived]` | Compute new columns from model outputs (per-row, aggregates, integrals) |
| `[output]` | Echo individual PK parameters or covariates by name |

These blocks are evaluated after the FOCE/FOCEI/SAEM optimiser has converged.
The resulting columns appear at the end of `fit$sdtab` in declaration order.

### Columns included automatically

Even without a `[derived]` or `[output]` block, the sdtab now includes:

- **ETA columns** - one per random effect (e.g. `ETA_CL`, `ETA_V`).
- **TAFD** - time after first dose for each observation record.
- **TAD** - time after last dose (steady-state-aware).

---

## Model file syntax

### `[derived]` block

Each line declares one derived column. Three forms are supported:

```
[derived]
  # 1. Per-row expression: evaluated independently at each observation
  KE     = CL / V
  T_HALF = 0.6931472 / KE      # can reference earlier derived names

  # 2. Aggregation: scalar per subject, repeated across all rows
  CMAX   = max(IPRED)                   # subject-level Cmax
  CMIN   = min(IPRED, TAD < 1e-10)     # trough (optional filter)
  TMAX   = tmax(IPRED)                  # TIME at peak IPRED

  # 3. Numeric integration: trapezoidal over a time window
  AUC_0_72 = integral(IPRED, from=0, to=72)          # explicit window
  AUC_TAU  = integral(IPRED, window=24, anchor=0)     # periodic (per-dose)
  AUC_DV   = integral(DV, from=0, to=72)             # data-based (obs times)
```

### `[output]` block

Lists column names to echo from the model, separated by whitespace:

```
[output]
  CL V KA       # individual PK parameters
```

Everything in `[derived]` is already included automatically - you do not
need to repeat derived names in `[output]`.

---

## Variables available in `[derived]` expressions

| Name | Meaning |
|---|---|
| `TIME` | Observation time from the dataset |
| `TAFD` | Time after first dose |
| `TAD` | Time after last dose (SS-aware) |
| `IPRED` | Individual prediction |
| `PRED` | Population prediction |
| `DV` | Observed value |
| `MACHEPS` | Machine epsilon (~2.22e-16); use `MACHEPS * 100` for tolerances |
| `CL`, `V`, `KA`, ... | Individual PK parameters |
| `ETA_CL`, `ETA_V`, ... | Empirical Bayes ETAs |
| `TVCL`, `TVV`, ... | Population (theta) parameters |
| Covariate names | Per-observation covariate values |
| Earlier derived names | Sequential evaluation: `T_HALF = 0.693 / KE` |

Arithmetic operators: `+`, `-`, `*`, `/`, `mod` (Euclidean remainder).
Comparison operators: `<`, `>`, `<=`, `>=`, `==`, `!=`.
Logical operators: `&&` (and), `||` (or), `!` (not).
Functions: `exp`, `log`, `sqrt`, `abs`, `floor`, `ceil`, `round`.
Scientific notation works: `1e-10`, `1.5E+3`.

Note: use `&&` not `and` for logical AND in filter conditions.

---

## Worked example: warfarin one-compartment

The bundled `warfarin_derived` example adds KE, T_HALF, CMAX, TMAX, AUC,
and periodic AUC to the standard warfarin model:

```{r load-model, eval = FALSE}
ex  <- ferx_example("warfarin_derived")
# ex$data resolves to the bundled warfarin dataset via the data alias.
fit <- ferx_fit(ex$model, ex$data, verbose = FALSE)
```

After fitting, `fit$sdtab` contains the standard diagnostic columns plus
all the derived columns declared in the model file:

```{r show-sdtab, eval = FALSE}
head(fit$sdtab[, c("ID", "TIME", "TAFD", "TAD", "IPRED",
                   "KE", "T_HALF", "CMAX", "AUC_0_72")])
```

```
  ID TIME  TAFD   TAD   IPRED     KE T_HALF    CMAX AUC_0_72
1  1  0.0   0.0  0.00  0.0000 0.0166  41.70  8.2341  326.1
2  1  0.5   0.5  0.50  2.7843 0.0166  41.70  8.2341  326.1
3  1  1.0   1.0  1.00  4.8261 0.0166  41.70  8.2341  326.1
4  1  2.0   2.0  2.00  7.1834 0.0166  41.70  8.2341  326.1
```

The `CMAX` and `AUC_0_72` columns are subject-level aggregates and are
therefore constant across all observation rows for a given subject.
`KE` and `T_HALF` vary by row because they are per-row expressions
(they depend on individual PK parameters which are subject-constant in
this model, so in practice they are also constant per subject here, but
per-row evaluation supports models where covariates change over time).

---

## Aggregation functions

### `max`, `min`, `tmax` with optional filter

The filter is a boolean expression evaluated at each observation. Only
rows where the filter is `true` contribute to the aggregate:

```
[derived]
  CMAX_D1  = max(IPRED, TAFD < 24)          # Cmax on day 1
  CMIN_TAU = min(IPRED, TAD < 1e-10)        # trough concentration
  TMAX     = tmax(IPRED)                    # time of overall Cmax
```

The scalar result is repeated for all observation rows of the subject.
When no row matches the filter, the result is `NaN`.

---

## Numeric integration

### Explicit window

```
AUC_0_72 = integral(IPRED, from=0, to=72)
```

By default, ferx evaluates the integrand on a fine internal grid
(500 steps) using the model's analytical PK prediction function.
This gives accurate integrals even for sparse sampling designs.

To use a coarser grid (faster for ODE models):

```
AUC_0_72 = integral(IPRED, from=0, to=72, step=0.5)
```

### DV-based integration

When the integrand references `DV`, the integration uses the actual
observation times (trapezoidal rule on the data):

```
AUC_DV_72 = integral(DV, from=0, to=72)
```

The `step=` keyword is ignored for DV-based integrals
(`W_DERIVED_STEP_IGNORED` is emitted).

### Periodic integration (per-dose window)

The `window=` form computes one integral per dosing interval and assigns
each observation the value for its window:

```
AUC_TAU = integral(IPRED, window=24, anchor=0, step=0.1)
```

With `anchor=0`, dosing windows are `[0, 24)`, `[24, 48)`, etc.

### Time-above-threshold

A constant integrand of `1.0` with a concentration filter gives hours
above a threshold:

```
TAM_TAU = integral(1.0, IPRED > MEC, window=24, anchor=0, step=0.1)
```

---

## TAFD and TAD in derived expressions

`TAFD` and `TAD` are available as first-class variables in `[derived]`
expressions without any declaration.

```
[derived]
  DAY   = floor(TAFD / 24) + 1    # dosing day (1-indexed)
  TAU_T = TAFD mod 24             # time within dosing interval
```

`TAD` is steady-state-aware: for dose rows with `SS=1` and `II > 0`,
the engine back-computes a virtual "last effective dose time" so that
`TAD` reflects the correct within-interval time at steady state.

For ADDL-expanded datasets (datasets with an `ADDL` column), each
expanded dose shifts the TAD anchor, so `TAD` resets near zero at each
dose time. Use `TAD < 1e-10` (not `TAD == 0`) to catch these reset
points robustly:

```
CMIN_TAU = min(IPRED, TAD < 1e-10)
```

---

## `[output]` block

Use `[output]` to echo individual PK parameter values or covariates
that are not already in sdtab:

```
[output]
  CL V KA
```

The names must be either individual parameter names (declared in
`[individual_parameters]`) or covariate column names from the dataset.
Using a name that is already in the mandatory sdtab minimum (e.g.
`TAFD`, `TAD`, or any ETA name) emits `W_OUTPUT_DUPLICATE`; the column
is still written but you do not need to declare it.

---

## Diagnostic codes

| Code | Severity | Meaning |
|---|---|---|
| `E_DERIVED_NAME_CONFLICT` | error | `[derived]` name clashes with a built-in column, theta, eta, or indiv-param name |
| `E_OUTPUT_UNKNOWN_COLUMN` | error | Name in `[output]` not recognised |
| `W_OUTPUT_DUPLICATE` | warning | Name in `[output]` already in the mandatory sdtab minimum |
| `W_DERIVED_COVARIATE_SHADOW` | warning | `[derived]` name shadows a covariate (allowed but may confuse) |
| `W_DERIVED_STEP_IGNORED` | warning | `step=` given for a DV-based integral; ignored |
| `W_ADDL_MISSING_II` | warning | ADDL > 0 but II is zero; additional doses not expanded |

---

## Post-fit analysis: Cmax vs. weight

After fitting a model with `[derived]` and `[output]`, you can join the
derived sdtab to any external data for downstream analysis. Here is a
sketch using ggplot2:

```{r ggplot-sketch, eval = FALSE}
library(ggplot2)

# One row per subject (CMAX is subject-constant; WT comes from [output])
subj <- fit$sdtab[!duplicated(fit$sdtab$ID), ]

ggplot(subj, aes(WT, CMAX)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Body weight (kg)", y = "Predicted Cmax (mg/L)")
```

---

## AUC and Cmax summary table

```{r summary-table, eval = FALSE}
# One row per subject: AUC and Cmax from [derived]
pk_summary <- aggregate(
  cbind(AUC_0_72, CMAX, T_HALF) ~ ID,
  data    = fit$sdtab,
  FUN     = function(x) x[[1L]]  # all rows equal for aggregates
)
print(pk_summary)
```

---

## Using ETAs from sdtab

ETA columns are now included automatically in sdtab, so you can use them
directly without reaching into `fit$ebe_etas`:

```{r eta-example, eval = FALSE}
# Shrinkage-aware correlation: ETA_CL vs. body weight
plot(fit$sdtab$WT, fit$sdtab$ETA_CL,
     xlab = "WT (kg)", ylab = "ETA_CL",
     main = "ETA_CL vs. body weight")
abline(h = 0, lty = 2)
```
