---
title: "Editing ferx model files programmatically"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Editing ferx model files programmatically}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ferx)
```

ferx model files (`.ferx`) are plain text, so you can create and modify them
entirely from R without opening an editor. This vignette covers the three
functions that support that workflow: `ferx_model()` (scaffold mode),
`ferx_model_get_section()`, and `ferx_model_set_section()`.

## Starting from scratch

`ferx_model(template = ...)` writes a skeleton `.ferx` file for one of five
built-in templates - `"1cpt_oral"` (default), `"1cpt_iv"`, `"2cpt_oral"`,
`"2cpt_iv"`, `"ode"` - and returns a `ferx_model` object that pipes straight
into `ferx_fit()`.

Pass `print = TRUE` to preview a template in the console without writing any
file — useful when you are exploring or copy-pasting a starting point:

```{r new-print}
ferx_model(print = TRUE)
```

```{r new-print-2cpt, eval = FALSE}
ferx_model(template = "2cpt_iv", print = TRUE)
```

To write a file, supply a `path`. By default the file is also opened in your
editor (`edit = TRUE`); pass `edit = FALSE` to skip that:

```{r new-write, eval = FALSE}
ferx_model(template = "1cpt_oral", path = "my_model.ferx", edit = FALSE)
```

## Inspecting a section

`ferx_model_get_section()` extracts the body of a named section and prints it.
Start with the bundled warfarin example:

```{r section}
ex <- ferx_example("warfarin")
ferx_model_get_section(ex$model, "parameters")
```

The return value is the character vector of lines, returned invisibly, so you
can capture it:

```{r section-capture}
params <- ferx_model_get_section(ex$model, "parameters")
params
```

## Scripted read-modify-write

The extract → modify → write-back pattern lets you change a model
programmatically without an editor. Here we raise the initial CL estimate in
the `parameters` section and write the result to a temporary copy:

```{r read-modify-write}
# work in a fresh temp file so re-knitting the vignette never reuses stale state
model_path <- tempfile(fileext = ".ferx")
stopifnot(file.copy(ex$model, model_path))

# read the section
lines <- ferx_model_get_section(model_path, "parameters")

# modify: change the TVCL initial value from 0.134 to 0.5
lines <- sub("TVCL\\([^,]+,", "TVCL(0.5,", lines)

# write back
ferx_model_set_section(model_path, "parameters", lines)

# verify
ferx_model_get_section(model_path, "parameters")
```

`ferx_model_set_section()` replaces only the targeted section; all other
sections are left untouched.

## Console-only workflow: create, tweak, fit

You can go from nothing to an estimated model without ever opening a text
editor. The steps below create a one-compartment oral model file, switch the
estimation method to FOCEI, and run `ferx_fit()`:

```{r console-only, eval = FALSE}
model_path <- file.path(tempdir(), "run1.ferx")

# 1. Write a skeleton
ferx_model(template = "1cpt_oral", path = model_path, edit = FALSE)

# 2. Switch method to focei
ferx_model_set_section(model_path, "fit_options", c(
  "  method     = focei",
  "  maxiter    = 300",
  "  covariance = true"
))

# 3. Fit (uses the bundled warfarin data for illustration)
ex   <- ferx_example("warfarin")
fit  <- ferx_fit(model_path, ex$data)
print(fit)
```

## Opening a model in an editor

`ferx_model_edit()` copies a bundled example to a destination directory (or
leaves a user-owned file in place) and opens it in your editor. It is the
interactive companion to the scripted functions above.

```{r model-edit, eval = FALSE}
# Copy the bundled warfarin model to tempdir() and open it
my_model <- ferx_model_edit(ex$model, dest = tempdir())

# After editing, save the result under a new name for version control
ferx_model_edit("run1.ferx", save_as = "run2.ferx")
```

The `dest` argument controls where the copy lands. If the source file is
already outside the installed package directory, `ferx_model_edit()` edits it
in place (no copy) unless `dest` is set explicitly.

## Validating a model file

`ferx_model_validate()` checks that all required sections are present and
reports the result. Use it after creating or editing a model to catch missing
sections before running the optimizer:

```{r validate-ok}
ferx_model_validate(ex$model)
```

A model missing required sections produces a clear error list:

```{r validate-bad}
bad <- tempfile(fileext = ".ferx")
writeLines(c(
  "[parameters]",
  "  theta TVCL(1.0, 0.001, 100.0)",
  "[structural_model]",
  "  pk one_cpt_oral(cl=CL, v=V, ka=KA)"
), bad)
ferx_model_validate(bad)
```

## Overwrite guards

By default `ferx_model(template = ...)` refuses to overwrite an existing file:

```{r overwrite-error}
model_path <- tempfile(fileext = ".ferx")
ferx_model(template = "1cpt_oral", path = model_path, edit = FALSE)  # creates it

tryCatch(
  ferx_model(template = "1cpt_oral", path = model_path, edit = FALSE), # would overwrite - error
  error = function(e) message(conditionMessage(e))
)
```

Opt in explicitly with `overwrite = TRUE`:

```{r overwrite-ok}
ferx_model(template = "1cpt_iv", path = model_path, overwrite = TRUE, edit = FALSE)
ferx_model_get_section(model_path, "structural_model")
```
