---
title: "Simulation, prediction, and uncertainty propagation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation, prediction, and uncertainty propagation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ferx)
```

## Overview

After a model has been fit, three functions cover the standard
post-estimation workflow:

| Function | Returns | Captures |
|---|---|---|
| `ferx_predict()` | population predictions (`PRED`) | typical-value trajectory (eta = 0) |
| `ferx_simulate()` | replicates of `IPRED` and `DV_SIM` | between-subject variability + residual error |
| `ferx_simulate_with_uncertainty()` | same + parameter draws | also propagates parameter uncertainty |

A fourth function, `ferx_simulate_adaptive()`, is covered at the end: it is not
a post-estimation summary but a forward simulation whose *dosing regimen is
decided at run time* by a controller in the model file - the feedback / dose-
titration path.

The three summary functions above all take the same `(model, data)` pair as
`ferx_fit()` and accept an
optional `fit =` argument that substitutes the fitted parameter values for the
model file's initial estimates. When `fit` is omitted, the initial values from
the `.ferx` file are used - which is occasionally useful for sanity-checking a
model before fitting, but the post-fit case is the common one.

---

## Setup

```{r setup-fit, message = FALSE, warning = FALSE}
ex  <- ferx_example("warfarin")
fit <- ferx_fit(ex$model, ex$data,
                method = "gn", covariance = TRUE, verbose = FALSE)
```

The examples below run live against this fit, so the numeric outputs stay in
sync with the package and the bundled warfarin dataset.

This chunk uses `message = FALSE, warning = FALSE` to keep the rendered
output tidy. The same fit run interactively prints a model-file
method-override warning (the bundled model file specifies `method = foce`
while the call overrides to `gn`) and a few diagnostic notes; both are
expected.

---

## Population prediction

`ferx_predict()` returns one row per observation in the data, with the
typical-value prediction (eta = 0) for each subject's covariates and dosing
record:

```{r predict}
pred <- ferx_predict(ex$model, ex$data, fit = fit)
head(pred)
```

Without `fit =`, the `[parameters]` block's initial estimates are used. The
columns are always `ID, TIME, PRED`.

---

## Individual simulation

`ferx_simulate()` draws fresh etas and epsilons and returns simulated
observations for each replicate. Each row carries:

| Column | Meaning |
|---|---|
| `DRAW`  | Always `1` for `ferx_simulate()`; present so downstream code can share aggregation logic with `ferx_simulate_with_uncertainty()` |
| `SIM`   | Replicate index, 1..`n_sim` |
| `ID`    | Subject ID from the data file |
| `TIME`  | Observation time from the data file |
| `IPRED` | Individual prediction (using the drawn eta) |
| `DV_SIM`| Simulated observation (IPRED plus residual error) |

```{r simulate}
sim <- ferx_simulate(ex$model, ex$data, n_sim = 3L, seed = 1L, fit = fit)
head(sim)
nrow(sim)            # n_sim * n_obs
unique(sim$SIM)
```

`seed` makes the draws reproducible across runs.

### Quick visual prediction interval

A common diagnostic is a per-time prediction interval over the simulated
replicates. With `n_sim = 100` and base aggregation:

```{r vpc}
sim <- ferx_simulate(ex$model, ex$data, n_sim = 100L, seed = 1L, fit = fit)

pi <- aggregate(DV_SIM ~ TIME, data = sim,
                FUN = function(x) quantile(x, c(0.05, 0.5, 0.95)))
head(pi)
```

To draw a visual predictive check with ggplot2 you would unpack the matrix
column produced by `quantile()` into named columns and layer the ribbon over
the observed `DV`:

```{r vpc-plot, eval = FALSE}
library(ggplot2)

q <- as.data.frame(pi$DV_SIM)
names(q) <- c("lo", "med", "hi")
pi_df <- data.frame(TIME = pi$TIME, q)

obs <- read.csv(ex$data)
obs <- obs[obs$EVID == 0 & obs$MDV == 0, ]
# DV is character in read.csv because dose rows use "." as a placeholder.
# Coerce after filtering observation rows so the ribbon plot gets a numeric axis.
obs$DV <- as.numeric(obs$DV)

ggplot() +
  geom_ribbon(data = pi_df, aes(TIME, ymin = lo, ymax = hi),
              fill = "steelblue", alpha = 0.25) +
  geom_line(data   = pi_df, aes(TIME, med),
            colour = "steelblue") +
  geom_point(data  = obs, aes(TIME, DV),
             alpha = 0.4) +
  labs(title = "Simulation-based 90% prediction interval",
       y     = "DV")
```

The simple aggregation above stratifies only by `TIME`. For a "real" VPC you
would also stratify by dose or covariate group, and bin times for sparse
designs - both are downstream data-frame manipulations and are not built into
ferx.

---

## Propagating parameter uncertainty

`ferx_simulate_with_uncertainty()` adds an outer loop: for each of
`n_uncertainty_draws` parameter sets drawn from the fit's uncertainty
distribution, `n_sim_per_draw` eta/eps replicates are simulated. The result
includes both the within-replicate variability and the parameter uncertainty
itself.

Two uncertainty sources are supported:

| `method` | Requires | Sampling |
|---|---|---|
| `"asymptotic"` (default) | `fit$cov_matrix` (i.e. `ferx_fit(..., covariance = TRUE)`) | Multivariate-normal draws around the ML estimate, using the fit's covariance matrix as the proposal. Internally the engine works in a packed (log-theta, Cholesky-omega, log-sigma) parameterisation so the draws respect positivity and PSD constraints. |
| `"sir"` | `fit$sir_resamples` (i.e. `ferx_fit(..., sir = TRUE, settings = list(sir_keep_samples = TRUE))`) | Resampling with replacement from the stored SIR resamples - importance-weighted draws that better reflect the true likelihood when the asymptotic approximation is poor. |

The return value has the same column set as `ferx_simulate()`, but `DRAW` now
ranges over 1..`n_uncertainty_draws`:

```{r sim-unc}
sims <- ferx_simulate_with_uncertainty(
  ex$model, ex$data, fit,
  n_uncertainty_draws = 5L,
  n_sim_per_draw      = 2L,
  seed                = 1L
)
head(sims)
nrow(sims)            # n_uncertainty_draws * n_sim_per_draw * n_obs
```

The row count is `n_uncertainty_draws * n_sim_per_draw * n_obs`. For an
uncertainty-aware prediction interval, aggregate over all rows at each `TIME`:

```{r sim-unc-pi}
pi_unc <- aggregate(DV_SIM ~ TIME, data = sims,
                    FUN = function(x) quantile(x, c(0.05, 0.5, 0.95)))
head(pi_unc)
```

The asymptotic method is fast but assumes the parameter likelihood is locally
quadratic around the ML estimate. For models with strong non-normality or
boundary issues, prefer the SIR method - but only when the cost of running SIR
at fit time is acceptable.

---

## State-reactive (adaptive) dosing

The three functions above replay a *fixed* regimen: the doses come from the
data and never change in response to the simulated state. `ferx_simulate_adaptive()`
is different - the regimen is decided at run time by a declarative controller in
the model file's `[adaptive_dosing]` block, which reads the simulated (optionally
assay-noised) state at each decision time and titrates the next dose. This is the
therapeutic-drug-monitoring / model-informed-precision-dosing path.

The data may still carry a **pre-scheduled base regimen** - ordinary dose rows,
such as a loading dose - and the controller *augments* it. The bundled
`adaptive_vanco_loading` example does exactly this: a 1500 mg vancomycin loading
dose rides on the data, and the controller titrates the daily maintenance dose to
hold the trough in a 10-15 mg/L window.

```{r adaptive-run, message = FALSE, warning = FALSE}
ex  <- ferx_example("adaptive_vanco_loading")
res <- ferx_simulate_adaptive(ex$model, ex$data, n_sim = 10L, seed = 1L)
str(res, max.level = 1)
```

`ferx_simulate_adaptive()` returns four data frames rather than one: the
per-observation `trajectories`, the realized dose `ledger` (`doses`), the
per-decision log (`decisions`, including holds and stops), and per-subject outcome
`metrics`. The controller's day-by-day view - the trough it observed and the rule
it fired - is in `decisions`:

```{r adaptive-decisions}
res$decisions[res$decisions$SIM == 1, c("DECISION", "TIME", "SIGNAL", "OUTCOME")]
```

The trough starts sub-therapeutic (the decayed loading dose, ~5.6 mg/L on day 1)
and climbs into the target window as the maintenance dose is titrated up.

One caveat worth stating explicitly: the dose **ledger holds controller-issued
doses only**. A pre-scheduled base dose - here the 1500 mg loading dose - never
appears in `doses`, and is excluded from `metrics$CUM_DOSE`. It *is* reflected in
the trajectories and the attainment metric. To recover the full realized regimen,
add the base doses from the input data back in:

```{r adaptive-ledger}
loading_dose   <- 1500
cum_controller <- res$metrics$CUM_DOSE[res$metrics$SIM == 1]
c(controller_only = cum_controller,
  full_realized   = cum_controller + loading_dose,
  loading_in_ledger = any(res$doses$AMT == loading_dose))
```

---

## Choosing between the four

- **`ferx_predict()`** when you only need the typical-value trajectory - for a
  PRED-vs-time overlay, or for forecasting under a new dosing schedule.
- **`ferx_simulate()`** when you want a prediction interval that reflects
  between-subject variability and residual error, but you are willing to treat
  the ML estimates as fixed.
- **`ferx_simulate_with_uncertainty()`** when the question is about uncertainty
  in the parameter values themselves: dose-recommendation intervals, model
  comparison via predictive performance, or any decision-relevant interval
  that would understate spread if parameter uncertainty were ignored.
- **`ferx_simulate_adaptive()`** when the regimen is not fixed but decided by
  feedback - dose titration to a target trough or AUC, with the rules living in
  the model file's `[adaptive_dosing]` block rather than in the data.
