---
title: "Special features: IOV, BLOQ, steady state, and unit scaling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Special features: IOV, BLOQ, steady state, and unit scaling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(ferx)
```

## Overview

This vignette is a tour of four `.ferx` features that go beyond the standard
single-dose oral PK example. Each section uses a bundled example model so you
can reproduce it without writing a model file from scratch.

| Feature | Bundled example | Key DSL/data element |
|---|---|---|
| Inter-occasion variability (IOV) | `warfarin_iov` | `kappa` declaration + `iov_column = OCC` |
| Below lower limit of quantification (BLOQ) | `warfarin_bloq` | `CENS` data column + `bloq_method = "m3"` |
| Steady-state dosing | `warfarin_ss` | `SS` and `II` data columns |
| Unit conversion / observation scaling | `warfarin_scaled` | `[scaling]` block with `obs_scale` |

All four are independent of one another - mix and match as needed.

---

## Inter-occasion variability

IOV captures variability that recurs within a single subject across distinct
occasions (visits, treatment cycles, ...). The bundled `warfarin_iov` model
declares an IOV term on clearance using the `kappa` keyword:

```{r iov-inspect}
ex_iov <- ferx_example("warfarin_iov")
ferx_model_inspect(ex_iov$model)
```

Three pieces have to line up:

1. **`[parameters]`** declares one or more `kappa` parameters with their
   omega-IOV variance, e.g. `kappa KAPPA_CL ~ 0.04`.
2. **`[individual_parameters]`** uses the kappa additively on the log scale,
   alongside the usual IIV ETA:
   ```
   CL = TVCL * exp(ETA_CL + KAPPA_CL)
   ```
3. **`[fit_options]`** points the engine at the occasion column:
   ```
   iov_column = OCC
   ```
4. **The data** carries an integer `OCC` column. Each value within a subject
   identifies an occasion; the engine draws one kappa per occasion per
   subject.

After fitting, `fit$omega_iov` is a square matrix on the kappa diagonal, and
`fit$estimates` reports each kappa as a row with `transform = "variance"`
(same convention as omega rows):

```{r iov-fit, eval = FALSE}
fit_iov <- ferx_fit(ex_iov$model, ex_iov$data, method = "focei")
fit_iov$omega_iov
#>            KAPPA_CL
#> KAPPA_CL 0.04679037

fit_iov$estimates
#>     param    transform   estimate se rse_pct lower_95 upper_95 estimate_natural lower_95_natural upper_95_natural init_as_sd
#>      TVCL     identity 0.17003443 NA      NA       NA       NA               NA               NA               NA      FALSE
#>       TVV     identity 8.62276910 NA      NA       NA       NA               NA               NA               NA      FALSE
#>      TVKA     identity 1.16618408 NA      NA       NA       NA               NA               NA               NA      FALSE
#>    ETA_CL     variance 0.03849065 NA      NA       NA       NA               NA               NA               NA      FALSE
#>     ETA_V     variance 0.01227511 NA      NA       NA       NA               NA               NA               NA      FALSE
#>    ETA_KA     variance 0.06600358 NA      NA       NA       NA               NA               NA               NA      FALSE
#>  PROP_ERR proportional 0.17643369 NA      NA       NA       NA               NA               NA               NA       TRUE
#>  KAPPA_CL     variance 0.04679037 NA      NA       NA       NA               NA               NA               NA      FALSE
# Output captured from a real focei fit with covariance = FALSE, hence
# the NA SE/CI columns. KAPPA_CL appears as the last row, alongside
# the other variance-scale entries. Kept eval = FALSE because the
# focei fit takes ~38 s on this example - too slow for the vignette
# build.
```

The IOV print labels follow the bare-name convention: `KAPPA_CL` is the
declared name; an unnamed declaration would fall back to `KAPPA1`,
`KAPPA2`, etc.

---

## Below lower limit of quantification (BLOQ)

When some observations sit below the assay's lower limit, the standard
Gaussian likelihood is biased. ferx implements Beal's M3 likelihood, which
treats each BLOQ observation as a left-censored draw and integrates over the
unobserved range.

Two pieces are needed:

1. **A `CENS` column in the data**: `1` for BLOQ (below LLOQ) rows, `-1` for rows right-censored above the ULOQ, `0` otherwise. The `DV`
   cell on a BLOQ row should carry the LLOQ value itself.
2. **Tell the engine to use M3**: either pass `bloq_method = "m3"` to
   `ferx_fit()`, or set `bloq_method = m3` in `[fit_options]`. The other
   choice is `bloq_method = "drop"`, which disables M3 but **keeps** the
   censored rows, fitting each at the limit value in `DV` as though it had been
   measured there. That biases the fit, and is rarely what the name suggests --
   to genuinely exclude the rows, pass `ignore = "CENS == 1"` instead.

```{r bloq-inspect}
ex_bloq <- ferx_example("warfarin_bloq")
ferx_model_inspect(ex_bloq$model)
```

The bundled `warfarin_bloq` model already sets `bloq_method = m3` in
`[fit_options]`. Inspect the data to see the `CENS` column:

```{r bloq-data}
data_bloq <- read.csv(ex_bloq$data)
head(data_bloq[data_bloq$CENS == 1, ], 3)
```

Running the fit produces the standard `ferx_fit` result; the M3 likelihood
contribution from BLOQ rows is folded into the OFV:

```{r bloq-fit, eval = FALSE}
fit_bloq <- ferx_fit(ex_bloq$model, ex_bloq$data, method = "focei")
```

The call-time argument overrides the model-file setting if both are present.
Pass `bloq_method = "m3"` explicitly when you want the choice to be visible
in the script.

---

## Steady-state dosing

Steady state lets you specify a maintenance dosing regimen without listing
every single dose in the data. The engine analytically computes the
steady-state initial condition for the dosing compartment given the dose, the
interval, and the current individual parameters - no model changes are
required.

The bundled `warfarin_ss` example dosing record carries:

```{r ss-data}
ex_ss <- ferx_example("warfarin_ss")
data_ss <- read.csv(ex_ss$data)
head(data_ss[data_ss$EVID == 1, ], 3)
```

Two extra columns make the row a steady-state dose:

| Column | Meaning |
|---|---|
| `SS = 1`  | Treat this row as a steady-state dose record |
| `II = 24` | Inter-dose interval (here 24 h, i.e. once-daily dosing) |

Subsequent observations are evaluated against the steady-state initial
condition. The model file itself is unchanged from the single-dose warfarin
example - steady state is entirely a data-side feature:

```{r ss-inspect}
ferx_model_inspect(ex_ss$model)
```

For the bundled steady-state warfarin example (tau = 24 h, half-life
~ 42 h - i.e. ke = CL/V ~ 0.0165/h), accumulation is about three-fold, so the
`DV` values sit in the 20-35 mg/L range vs. 5-14 mg/L for the single-dose
version.

```{r ss-fit, eval = FALSE}
fit_ss <- ferx_fit(ex_ss$model, ex_ss$data, method = "focei")
```

`SS` and `II` interact correctly with covariates and IIV: each subject's
steady-state initial condition is computed from their own individual
parameters.

---

## Unit conversion and observation scaling

The `[scaling]` block converts model predictions before they are compared to
`DV`. Use it when the model is parameterised in one unit but the data are
recorded in another - for example, dose in micrograms but observations in
mg/L, or a central-compartment ODE in amounts but observations in
concentrations.

The bundled `warfarin_scaled` example records dose in micrograms
(`AMT = 100000` ug vs the usual `100` mg) and observations in mg/L. The
model adds a single line:

```
[scaling]
  obs_scale = 1000
```

This divides every prediction by 1000 before the residual is computed, so the
sigma is expressed on the mg/L scale.

```{r scaled-inspect}
ex_scaled <- ferx_example("warfarin_scaled")
ferx_model_inspect(ex_scaled$model)
```

Three forms of `obs_scale` are supported (see `?ferx_fit` for the full
section on the `[scaling]` block):

| Form | Syntax | Use case |
|---|---|---|
| Constant divisor | `obs_scale = 1000` | Pure unit conversion |
| Expression | `obs_scale = V` | Volume-of-distribution scaling (concentration = amount / V), referencing thetas / etas / individual parameters |
| Per-compartment | `obs_scale[CMT=1] = 1000` | Multi-endpoint models with different units per observation compartment |

A more general Form C lets you write the readout expression directly
(`y = central / V`), useful for non-trivial transformations.

A *uniform* expression divisor (form B) or ODE readout (form C) is
differentiated **exactly** under the default `gradient = auto` (ferx-core #486),
so leave the gradient alone -- form B in particular runs on the analytical
closed-form path, not through the ODE machinery. Only the per-CMT variants
(`obs_scale[CMT=N]`, `y[CMT=N]`) fall back to finite differences, and they do so
**silently**: there is no error and no warning, so check `fit$gradient_used` if
it matters.

`[scaling]` is not yet supported on SDE (`[diffusion]`) models.

```{r scaled-fit, eval = FALSE}
fit_scaled <- ferx_fit(ex_scaled$model, ex_scaled$data, method = "focei")
fit_scaled$theta
# Should converge to the same parameter values as the standard warfarin
# fit because only the units differ.
```

---

## Combining features

These four features are independent and can be combined in the same model. A
study with once-daily steady-state dosing, BLOQ observations, occasion-level
CL variability, and concentration units derived via a volume-scaled readout
would carry all of:

- a `kappa KAPPA_CL ~ 0.04` declaration and an `iov_column = OCC` setting,
- a `CENS` column and `bloq_method = m3` setting,
- `SS = 1` and `II = 24` columns on the dosing rows,
- a `[scaling]` block with `obs_scale = V` (Form B) and an explicit
  `gradient = fd` in `[fit_options]`.

No bundled `.ferx` exercises all four at once; build one by concatenating the
relevant sections from the per-feature examples above. Validate the result
with `ferx_model_validate()` before fitting.

Each feature is documented in `?ferx_fit` along with its data-side
requirements; this vignette is the narrative tour.
